---
title: "main"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
```

Get symbols from:

- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis' FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

# Cleaning
Convert all XTS objects to tibbles.
```{r}
dgs10 <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
)
amd <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
)
phi <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
)
vix <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
)
eth <- tibble(
  date = index(`ETH-USD`),
  open = as.numeric(`ETH-USD`$`ETH-USD.Open`),
  high = as.numeric(`ETH-USD`$`ETH-USD.High`),
  low = as.numeric(`ETH-USD`$`ETH-USD.Low`),
  close = as.numeric(`ETH-USD`$`ETH-USD.Close`),
  volume = as.numeric(`ETH-USD`$`ETH-USD.Volume`),
  adjusted = as.numeric(`ETH-USD`$`ETH-USD.Adjusted`),
)
```