---
title: "FINSTAD Financial Database"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk, zoo)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
library(zoo)
```

### Comparing the different datasets

We first get symbols from the following sources:
- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis' FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("FEDFUNDS",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

Then we can look at the macroeconomic data,

The data is taken from the Federal Reserve Economic Database (FRED) from the Federal Reserve Bank of St. Louis, where we load in the _Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity_ (**DGS10**), _Consumer Price Index for All Urban Consumers_ (**CPIAUCSL**), and the _Federal Funds Effective Rate_ (**FEDFUNDS**).

**DGS10** is used as a proxy for the "risk-free rate" in finance, it helps investors gauge inflation expectations, economic growth, and monetary policy shifts. **CPIAUCSL** is seasonally adjusted to gauge monthly inflation metrics on consumer goods. **FEDFUNDS** is the monthly average interest rate at which banks lend reserves to each other overnight, and reflects the stance of U.S. monetary policy (tightening vs. easing cycles).

These pieces of macroeconomic data would help in understanding broader market cycles and how systemic risks affect stock valuations.
```{r}
head(DGS10)
head(CPIAUCSL)
head(FEDFUNDS)
```

We can then take a look at our chosen financial market data,

U.S. Asset - Advanced Micro Devices, Inc. (AMD)
Philippine Asset - PLDT, Inc.(PHI)
Cryptocurrency - Etherium (ETH or ETH-USD)
Required - State Street SPDR S&P 500 ETF Trust (SPY)

```{r}
ETH <- `ETH-USD`

head(AMD)
head(PHI)
head(ETH)
head(SPY)
```

Lastly, we take a look at the _Cboe Volatility Index_ (**VIX**), this is not an individual stock but a real time market index created by the Chicago Board Options Exchange (Cboe). This measures the market's expectation of 30-day volatility in the S&P 500 index.

This adds aid on top of macroeconomic data to further gauge systemic risks, VIX is known as the "Fear Gauge" which reflects levels of anxiety, complacency, or uncertainty in the market, when this "Fear Gauge" level rises it implies larger expected swings in stock prices.
```{r}
head(VIX)
```

# Cleaning

### 1. Converting XTS to tibbles
First we convert all XTS objects to tibbles for easier data manipulation.
```{r}
dgs10_tbl <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl_tbl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy_tbl <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
  adj_close = as.numeric(Ad(SPY)),
)
amd_tbl <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
  adj_close = as.numeric(Ad(AMD)),
)
phi_tbl <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
  adj_close = as.numeric(Ad(PHI)),
)
vix_tbl <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
  adj_close = as.numeric(Ad(VIX)),
)
eth_tbl <- tibble(
  date = index(ETH),
  open = as.numeric(ETH$`ETH-USD.Open`),
  high = as.numeric(ETH$`ETH-USD.High`),
  low = as.numeric(ETH$`ETH-USD.Low`),
  close = as.numeric(ETH$`ETH-USD.Close`),
  volume = as.numeric(ETH$`ETH-USD.Volume`),
  adjusted = as.numeric(ETH$`ETH-USD.Adjusted`),
  adj_close = as.numeric(Ad(ETH)),
)
```

### 2. Asset Identifier
We add an asset identifier per asset to see the data better in table structures.
```{r}
amd_tbl <- amd_tbl |> 
  mutate(asset = "AMD")

eth_tbl <- eth_tbl |> 
  mutate(asset = "ETH")

phi_tbl <- phi_tbl |> 
  mutate(asset = "PHI")

spy_tbl <- spy_tbl |> 
  mutate(asset = "SPY")

vix_tbl <- vix_tbl |> 
  mutate(asset = "VIX")
```

### 3. Computing daily returns

Daily returns are computed using the basic return formula
```{r}
amd_tbl <- amd_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

eth_tbl <- eth_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

phi_tbl <- phi_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

spy_tbl <- spy_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

vix_tbl <- vix_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )
```

### 4. Removing missing rows
Null rows are dropped to preserve data integrity
```{r}
amd_tbl <- amd_tbl  |> 
  filter(!is.na(return))

eth_tbl <- eth_tbl  |> 
  filter(!is.na(return))

phi_tbl <- phi_tbl |> 
  filter(!is.na(return))

spy_tbl <- spy_tbl  |> 
  filter(!is.na(return))

vix_tbl <- vix_tbl  |> 
  filter(!is.na(return))
```

### 5. Examine the data
```{r}
glimpse(amd_tbl)
glimpse(eth_tbl)
glimpse(phi_tbl)
glimpse(spy_tbl)
glimpse(vix_tbl)
```

Summary Tables
```{r}
summary(amd_tbl)
summary(eth_tbl)
summary(phi_tbl)
summary(spy_tbl)
summary(vix_tbl)
```

### 6. Preparing Macroeconomic variables
```{r}
treasury_tbl <- tibble(
  date = index(DGS10),
  treasury_10y = as.numeric(DGS10$DGS10)
)
```

```{r}
cpi_tbl <- tibble(
  date = index(CPIAUCSL),
  cpi = as.numeric(CPIAUCSL$CPIAUCSL)
)
```

```{r}
fedfunds_tbl <- tibble(
  date = index(FEDFUNDS),
  fedfunds = as.numeric(FEDFUNDS$FEDFUNDS)
)
```

```{r}
head(treasury_tbl)
head(cpi_tbl)
head(fedfunds_tbl)
```

### 7. Create an Observation Summary Table and saving it
We first count the observations
```{r}
treasury_tbl <- treasury_tbl  |> 
  filter(!is.na(treasury_10y))

cpi_tbl <- cpi_tbl  |> 
  filter(!is.na(cpi))

fedfunds_tbl <- fedfunds_tbl  |> 
  filter(!is.na(fedfunds))

nrow(amd_tbl)
nrow(eth_tbl)
nrow(phi_tbl)
nrow(spy_tbl)
nrow(vix_tbl)
nrow(treasury_tbl)
nrow(cpi_tbl)
nrow(fedfunds_tbl)
```

Creating the observation summary table
```{r}
obs_summary <- tibble(
  Dataset = c("AMD",
              "Ethereum",
              "PLDT",
              "SPY",
              "VIX",
              "Treasury",
              "CPI",
              "Fed Funds Rate"),
  Observations = c(
    nrow(amd_tbl),
    nrow(eth_tbl),
    nrow(phi_tbl),
    nrow(spy_tbl),
    nrow(vix_tbl),
    nrow(treasury_tbl),
    nrow(cpi_tbl),
    nrow(fedfunds_tbl)
  )
)

obs_summary
```

Saving the clean tables
```{r}
amd_clean <- amd_tbl
eth_clean <- eth_tbl
phi_clean <- phi_tbl
spy_clean <- spy_tbl
vix_clean <- vix_tbl
treasury_clean <- treasury_tbl
cpi_clean <- cpi_tbl
fedfunds_clean <- fedfunds_tbl
```

# Database Integration

After cleaning all the data, we collate the rows into 1 unified dataset through the usage of SQL joins. We first need to prepare the necessary tables for unifying the datasets.

## 1. Renaming variables
```{r}
amd_sql <- amd_clean  |> 
  select(date,
         amd_price = adj_close,
         amd_return = return)

eth_sql <- eth_clean  |> 
  select(date,
         eth_price = adj_close,
         eth_return = return)

phi_sql <- phi_clean  |> 
  select(date,
         phi_price = adj_close,
         phi_return = return)

spy_sql <- spy_clean  |> 
  select(date,
         spy_price = adj_close,
         spy_return = return)

vix_sql <- vix_clean |> 
  select(date,
         vix = adj_close,
         vix_return = return)
```

```{r}
head(amd_sql)
head(eth_sql)
head(phi_sql)
head(spy_sql)
head(vix_sql)
```

## 2. The Trading Calendar Problem

Our four core assets do not trade on the same schedule:

- **SPY** and **AMD** follow the NYSE/NASDAQ calendar (U.S. weekdays, minus U.S. market holidays).
- **PHI** (PLDT ADR) also trades on a U.S. exchange (NYSE), so it largely follows the same calendar as SPY/AMD, but can still diverge on days one market is closed and the other isn't (e.g. differing holiday schedules).
- **ETH** trades every single calendar day, including weekends and all holidays, since crypto markets never close.
- **VIX** follows the CBOE/NYSE calendar, similar to SPY.

This means a naive join on `date` will not line up perfectly — ETH alone contributes roughly 30% more calendar dates per year than the equity assets. This mismatch is exactly why the join strategy matters, and why we need to demonstrate multiple join types before picking one.

## 3. Demonstrating Join Types

Using `spy_sql` (equity calendar) and `eth_sql` (7-day calendar) as the clearest illustration of the mismatch:

```{r}
cat("SPY rows: ", nrow(spy_sql), "\n")
cat("ETH rows: ", nrow(eth_sql), "\n")
```

**`inner_join()`** — keeps only dates present in *both* datasets. This throws away every weekend/holiday ETH observation that SPY doesn't have.
```{r}
demo_inner <- spy_sql |> 
  inner_join(eth_sql, by = "date")

cat("inner_join rows: ", nrow(demo_inner), "\n")
```

**`full_join()`** — keeps every date present in *either* dataset, filling the missing side with `NA`. This preserves ETH's weekend rows but introduces `NA` for SPY/AMD/PHI on those same days, since equities simply didn't trade.
```{r}
demo_full <- spy_sql |> 
  full_join(eth_sql, by = "date")

cat("full_join rows: ", nrow(demo_full), "\n")
cat("Rows where SPY is NA (i.e. weekend/holiday-only ETH dates): ",
    sum(is.na(demo_full$spy_price)), "\n")
```

**`anti_join()`** — shows *only* the rows that don't have a match, which is useful for auditing exactly which dates get dropped or orphaned by a join. Here, it shows every date ETH traded that SPY didn't:
```{r}
demo_anti <- eth_sql |> 
  anti_join(spy_sql, by = "date")

cat("Dates ETH traded but SPY did not: ", nrow(demo_anti), "\n")
head(demo_anti)
```

**`left_join()`** — anchors on the left-hand table's dates and pulls in matches from the right, discarding right-only rows. Anchoring on `spy_sql` keeps every SPY trading day and attaches ETH's price on that date where available:
```{r}
demo_left <- spy_sql |> 
  left_join(eth_sql, by = "date")

cat("left_join rows: ", nrow(demo_left), "\n")
cat("Rows where ETH is NA: ", sum(is.na(demo_left$eth_price)), "\n")
```

## 4. Final Join Strategy

Our final database is meant to answer an **equity/benchmark-driven investment question** — comparing AMD, PHI, and ETH against SPY as the reference benchmark, alongside macro conditions. Given that framing:

- `inner_join()` is too aggressive — it would silently drop every day where even one asset didn't trade, shrinking the dataset unnecessarily and losing information about days SPY *did* move.
- `full_join()` keeps the most data, but introduces weekend/holiday rows where SPY, AMD, PHI, and VIX are all structurally `NA` (they simply have no observation those days) — this isn't "missing" data to be filled, it's a non-event, so those extra rows have no analytical value for an equity-anchored study.
- `anti_join()` is a diagnostic tool, not a database-building join — we use it above to confirm exactly how many ETH-only dates exist, but it doesn't belong in the final chain.
- **`left_join()` anchored on `spy_sql`** is the best choice, it preserves every SPY trading day (our benchmark and reference calendar), attaches whichever other series have data on that date, and cleanly produces `NA` only where a genuine data gap exists (e.g. ETH trading on a day AMD didn't, or a macro series not yet updated) rather than manufacturing rows that don't correspond to any equity trading activity.

## 5. Building the Unified Database
```{r}
unified_db <- spy_sql |> 
  left_join(amd_sql,
            by = "date") |> 
  left_join(phi_sql,
            by = "date") |>
  left_join(eth_sql,
            by = "date") |> 
  left_join(vix_sql,
            by = "date") |> 
  left_join(treasury_clean,
            by = "date") |> 
  left_join(cpi_clean,
            by = "date") |> 
  left_join(fedfunds_clean,
            by = "date") |> 
  arrange(date) |>
  tidyr::fill(treasury_10y,
              cpi,
              fedfunds,
              .direction = "downup")

glimpse(unified_db)
head(unified_db)
view(unified_db)
```

`treasury_10y` and `fedfunds` are only reported on U.S. business days, while `cpi` is only reported once a month. Since our final database is anchored on SPY's daily trading calendar, joining these lower-frequency series directly leaves `NA` on every date. We resolve this with `tidyr::fill(..., .direction = "downup")`, which carries the most recent known value forward through time (e.g. a CPI print from early June applies to every trading day until the next print in early July), and also back fills any leading rows at the very start of the sample that come before the first available reading. Using only `"down"` would leave those first few weeks of 2020 as `NA`, since there is nothing earlier to carry forward from. This is standard practice for macro series in daily financial datasets, since the "current" CPI or Fed Funds level doesn't change on the days between official releases, and a short leading gap is filled with the nearest known value rather than left blank.

## 6. Observation Tracking

Tracking row counts at each stage of the join chain shows exactly where (and why) the dataset shrinks or holds steady.
```{r}
obs_track_spy    <- spy_sql
obs_track_amd    <- obs_track_spy |> left_join(amd_sql, by = "date")
obs_track_phi    <- obs_track_amd |> left_join(phi_sql, by = "date")
obs_track_eth    <- obs_track_phi |> left_join(eth_sql, by = "date")
obs_track_vix    <- obs_track_eth |> left_join(vix_sql, by = "date")
obs_track_treas  <- obs_track_vix |> left_join(treasury_clean, by = "date")
obs_track_cpi    <- obs_track_treas |> left_join(cpi_clean, by = "date")
obs_track_ff     <- obs_track_cpi |> left_join(fedfunds_clean, by = "date")

join_tracking <- tibble(
  Step = c("Original SPY (anchor)",
           "+ AMD",
           "+ PHI",
           "+ ETH",
           "+ VIX",
           "+ Treasury (DGS10)",
           "+ CPI",
           "+ Fed Funds Rate"),
  Rows_After_Join = c(
    nrow(obs_track_spy),
    nrow(obs_track_amd),
    nrow(obs_track_phi),
    nrow(obs_track_eth),
    nrow(obs_track_vix),
    nrow(obs_track_treas),
    nrow(obs_track_cpi),
    nrow(obs_track_ff)
  )
)

join_tracking
```

Because every join in the chain is a `left_join()` anchored on `spy_sql`, **row count never shrinks past the original SPY row count** — a `left_join()` can only add columns/matches or introduce `NA`, never drop rows from the left-hand side. This is exactly the guarantee we wanted from our join strategy: the equity benchmark's trading calendar is preserved intact all the way through.

Checking null rows after the join and fill step:
```{r}
colSums(
  is.na(
    unified_db
  )
)
```

## 7. Dropping Null Rows

After `tidyr::fill(..., .direction = "downup")`, the macro columns (`treasury_10y`, `cpi`, `fedfunds`) should have zero remaining `NA`, including at the very start of the sample. Any `NA` still present at this point belongs to the **asset price/return columns** (`amd_price`, `amd_return`, `phi_price`, `phi_return`, `eth_price`, `eth_return`, `vix`, `vix_return`) — these represent real gaps where a given asset simply didn't trade on a day SPY did (e.g. a PSE holiday that isn't a NYSE holiday), not something we can reasonably forward-fill without fabricating price movement that never happened.

We drop these remaining rows, since imputing a stock/crypto price would distort the return and volatility statistics that our entire analysis depends on:
```{r}
unified_clean <- unified_db |> 
  tidyr::drop_na(amd_price, amd_return,
                 phi_price, phi_return,
                 eth_price, eth_return,
                 vix, vix_return,
                 spy_price, spy_return)

cat("Rows before dropping asset-level NA: ", nrow(unified_db), "\n")
cat("Rows after dropping asset-level NA: ", nrow(unified_clean), "\n")
cat("Rows dropped: ", nrow(unified_db) - nrow(unified_clean), "\n")

colSums(is.na(unified_clean))
```

`unified_clean` is our final analytical dataset, used for all descriptive statistics, SQL investigations, and visualizations going forward.

# SQL Investigation

Using the SQL-based query functions select(), filter(), mutate(), arrange(), group_by() and summarise(), we investigate 10 financial questions to broaden understanding of observed relationships within the dataset.

## Investigation 1: Biggest single-day moves per asset

**Objective:** Find the top 3 single-day returns for each of the four core assets to see which asset produces the largest short-term stocks.
```{r}
returns_long <- unified_clean |>
  select(date, amd_return, eth_return, phi_return, spy_return) |>
  tidyr::pivot_longer(cols = -date, names_to = "asset", values_to = "return")

top_gains <- returns_long |>
  group_by(asset) |>
  arrange(desc(return), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  filter(rank <= 3) |>
  select(asset, date, return, rank)

top_gains
```

Using SQL queries, we start with selecting the date and each asset’s return through a `select()` call, `group_by()` is then used to group the data by asset type (AMD, ETH, PHI SPY). The data is then arranged by largest to smallest returns through `arrange()`, then  coupled with a given rank per asset and presents the top 3 ranks through `mutate()` and `filter()`. 

The table reveals that AMD, ETH, and PHI all produced large single-day returns during their top-3 ranked days. AMD and ETH both exceeded +20% in a single trading session, with ETH's best day reaching +25.3% (2021-05-24) and AMD's reaching +23.8% (2025-04-09). SPY shows a much more modest top-3 range of roughly 8.5% to 10.5%, which is expected. These best days possibly represent the outlier days that drive a large share of an asset's total volatility. For AMD, the top days cluster around April–October 2025 (implies sector-specific successes like increased demand of chips), while for ETH, the top-3 dates span 2021 and 2024–2025, which are consistent with known periods of major crypto market rallies. The Philippine asset's top days are more spread out (2020, 2022, 2024) and its range is noticeably smaller than that of AMD's or ETH's suggesting it experiences less extreme single-day upturns.

## Investigation 2: Monthly average returns

**Objective:** Compute the average daily return per calendar month for each asset to identify which months or periods were consistently strong or weak.
```{r}
monthly_avg_returns <- unified_clean |>
  mutate(year_month = format(date, "%Y-%m")) |>
  group_by(year_month) |>
  summarise(
    avg_amd = mean(amd_return, na.rm = TRUE),
    avg_eth = mean(eth_return, na.rm = TRUE),
    avg_phi = mean(phi_return, na.rm = TRUE),
    avg_spy = mean(spy_return, na.rm = TRUE)
  )

monthly_avg_returns
```

Using the new unified dataset, we compute an average of the daily return to see which months are strong or weakpoints for each asset in our database. Using the `mutate(`) function, the dates are formatted from `YYYY-MM-DD` to `YYYY-MM` to prepare for the grouping of data by months and subsequent averaging of daily returns. Due to the amount of rows, we can simplify the viewing of the lowest and highest months per year per asset where we use `group_by()` per asset and return the best and worst performing months.

```{r}
monthly_long <- monthly_avg_returns |>
  tidyr::pivot_longer(
    cols = c(avg_amd, avg_eth, avg_phi, avg_spy),
    names_to = "asset",
    values_to = "avg_return"
  )

best_months <- monthly_long |>
  group_by(asset) |>
  slice_max(avg_return, n = 1) |>
  rename(best_month = year_month, best_return = avg_return)

worst_months <- monthly_long |>
  group_by(asset) |>
  slice_min(avg_return, n = 1) |>
  rename(worst_month = year_month, worst_return = avg_return)

best_worst_months <- best_months |>
  left_join(worst_months, by = "asset") |>
  select(asset, best_month, best_return, worst_month, worst_return)

best_worst_months
```
Then we find the best and worst months based on return value, where `group_by()` is used to group the values asset-wise and find the maximum or minimum return value in the column through finding the first value in a `slice_max()` and `slice_min()` call.

This table identifies an asset's worst and best months in terms of its average returns across the sample period. The best month column shows no shared patterns or timing across the assets (AMD: 2025-10, ETH: 2020-04, PHI: 2021-08, SPY: 2020-04), but a clear pattern emerges between the worst months across all assets. AMD (2022-05), SPY (2022-06), and ETH (2022-04) all posted their worst average monthly returns within a narrow three-month window in mid-2022, with losses ranging from -1.48% (AMD) to -2.49% (ETH). This clustering of downturns suggests a macroeconomic event over sector-specific events that possibly affected the performance of these stocks. This low performance period coincides with the 2022 stock market decline, due to an aggressive rate-hiking cycle by the U.S. Federal Reserve  to combat post-pandemic inflation, coinciding with the Russian invasion of Ukraine. PHI on the other hand showed a divergent pattern in its worst month (2022-12), suggesting it is not tightly coupled with U.S. monetary policy changes. 

## Investigation 3: SPY selloff days and cross-asset behavior

**Objective:** Identify SPY's 10 worst trading days and check how AMD, PHI, and ETH performed on those same dates, to test whether any of them behave as a diversifier during U.S. market selloffs.
```{r}
spy_selloffs <- unified_clean |>
  filter(spy_return < 0) |>
  arrange(spy_return) |>
  select(date, spy_return, amd_return, phi_return, eth_return) |>
  head(10)

spy_selloffs
```

A `filter()` call is first used to isolate only the days where `spy_return` was negative, restricting the dataset to actual down days rather than the full trading history. The remaining rows are then sorted from most severe to least severe using `arrange()` on `spy_return` in ascending order, so that SPY's largest single-day losses appear first. A `select()` call narrows the table down to the date and each asset's return (SPY, AMD, PHI, and ETH), allowing their behavior to be compared side by side on the same date. Finally, `head(10)` limits the output to SPY's ten worst trading days in the sample.

 The ten days in the sample represent SPY’s worst single-day returns, ranging from -4.5% to -10.9%. The worst dates cluster around historical market stress dates, like the crash during the pandemic in March of 2020 (7 of 10 observations) and tariffs from April of 2025 (3 of 10 observations). AMD shows the strongest correlation with SPY, as on every one of SPY’s worst dates, AMD has a negative return. This is consistent with AMD having a higher beta as a tech stock in the S&P 500.

## Investigation 4: High-VIX vs low-VIX return and volatility comparison

**Objective:** Split trading days into High VIX and Low VIX regimes (relative to the median VIX level) and compare average return and volatility in each regime.

```{r}
vix_median <- median(unified_clean$vix, na.rm = TRUE)

vix_regime_stats <- unified_clean |>
  mutate(vix_regime = if_else(vix > vix_median, "High VIX", "Low VIX")) |>
  group_by(vix_regime) |>
  summarise(
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    sd_spy_return = sd(spy_return, na.rm = TRUE),
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    sd_amd_return = sd(amd_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    sd_phi_return  = sd(phi_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    sd_eth_return  = sd(eth_return, na.rm = TRUE),
    n_days = n()
  )

vix_regime_stats
```
We classify each trading day into a volatility regime based on the VIX level and compare average returns and volatility across those regimes. The median VIX value across the full sample is first computed separately to serve as the split threshold. A `mutate()` call then creates a new `vix_regime` column, labeling each day as "High VIX" or "Low VIX" depending on whether that day's VIX reading falls above or below the sample median. `group_by()` is used to group the data by this newly created regime, and `summarise()` then computes the average return (`mean()`) and volatility (`sd()`) of both SPY and AMD returns within each group, along with a count of trading days (`n()`) falling into each regime. 

We use VIX to calculate market risk, or periods of low returns with high volatility, and if VIX patterns correlate with the average returns of an asset. As seen in the resulting table the high and low VIX instances are relatively equal, making this a fair assessment on how VIX is able to act as a "fear gauge". 'High VIX' consistently has a lower average of returns and low VIX being the opposite, with the asset's standard deviation also following the same pattern, larger volatility/swing in high VIX and lower in low VIX periods. This shows how VIX is able to accurately gauge market volatility.

## Investigation 5: Returns across Fed Funds rate regimes

**Objective:** Classifying each trading day by whether the Fed Funds Rate was rising, falling, or holding steady, then compare average returns across the three regimes to gauge rate sensitivity.

```{r}
fedfunds_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    fedfunds_change = fedfunds - lag(fedfunds),
    rate_regime = case_when(
      fedfunds_change > 0 ~ "Rate Hike",
      fedfunds_change < 0 ~ "Rate Cut",
      TRUE ~ "Held Steady"
    )
  ) |>
  filter(!is.na(rate_regime)) |>
  group_by(rate_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

fedfunds_trend
```

The data is first sorted chronologically using `arrange()` to ensure changes are calculated in proper date order. A `mutate()` call then computes `fedfunds_change` as the difference between each day's rate and the prior day's rate using `lag()`, and a `case_when()` statement classifies each day into a class "Rate Hike," "Rate Cut," or "Held Steady" based on the sign of that change. A `filter()` call removes the first row of the dataset, which has no prior value to compare against and would otherwise return a missing regime classification. `group_by()` is then used to group the data by rate regime, and `summarise()` computes the average return of AMD, ETH, PHI, and SPY within each group, along with a count of trading days (`n()`) in each regime.

In the resulting table, there is a clear imbalance between the number of days per rate class but this is expected as FEDFUNDS is a policy rate, not a continuously trading price, hence majority of days will naturally fall under the "Held Steady" class because rate changes are rare, discrete events.This also makes average returns sensitive to rate cuts and hikes, despite the small sample sizes for Rate Hike and Rate Cut days, a consistent pattern shows where every asset earns its highest average return during Rate Cut periods, and its lowest/near-lowest average during "Held Steady" periods. However, this pattern must be interpreted cautiously, as with only 8 Rate Cut days and 17 Rate Hike days in the entire sample, these averages are highly sensitive to one or two extreme trading days rather than reflecting a strong statistical relation.

## Investigation 6: Returns across inflation regimes

**Objective:** Classify each trading day by whether CPI was rising or falling/flat that month, then compare average returns across the two inflation regimes.

```{r}
cpi_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    cpi_change = cpi - lag(cpi),
    inflation_regime = if_else(cpi_change > 0, "CPI Rising", "CPI Falling or Flat")
  ) |>
  filter(!is.na(inflation_regime)) |>
  group_by(inflation_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

cpi_trend
```

The data is first sorted chronologically using arrange() so that changes are computed in the correct date sequence. A `mutate()` call then calculates `cpi_change` as the difference between each day's CPI value and the prior day's value using `lag()`, and an `if_else()` statement classifies each day as "CPI Rising" or "CPI Falling or Flat" depending on whether that change is positive. A `filter()` call removes the first row, which lacks a prior CPI value to compare against. `group_by()` groups the data by inflation regime, and `summarise()` computes the average return of AMD, ETH, PHI, and SPY within each group, along with the number of trading days (`n()`) falling into each classification. 

Looking at the resulting table, the number of days per class remains imbalanced, but this is expected as CPI is a monthly economic indicator and forward-filled onto the SPY trading calendar. Despite the small sample for "CPI Rising" a consistent pattern emerges with SPY, ETH, and PHI, where they all show noticeably higher average returns during CPI Rising periods compared to CPI Falling or Flat. AMD on the other hand shows a slightly lower average return during CPI Rising periods. This singular divergence may reflect higher sensitivity to sector-specific news.

## Investigation 7: Co-movement across all four assets

**Objective:** Count how often all four assets move in the same direction on the same day (all up or all down), to measure how correlated the portfolio really is versus how often the assets diverge.

```{r}
co_movement <- unified_clean |>
  mutate(
    all_positive = amd_return > 0 & eth_return > 0 & phi_return > 0 & spy_return > 0,
    all_negative = amd_return < 0 & eth_return < 0 & phi_return < 0 & spy_return < 0
  ) |>
  summarise(
    all_up_days = sum(all_positive, na.rm = TRUE),
    all_down_days = sum(all_negative, na.rm = TRUE),
    total_days = n(),
    pct_co_moving = (sum(all_positive, na.rm = TRUE) + sum(all_negative, na.rm = TRUE)) / n() * 100
  )

co_movement
```

A `mutate()` call creates two classes `all_positive` and `all_negative`, checking whether AMD, ETH, PHI, and SPY returns are simultaneously positive or simultaneously negative on a given day. `summarise()` is then used to aggregate these flags across the full dataset, a `sum()` call counts the number of all-up days and all-down days, `n()` returns the total number of trading days in the sample, and a combined calculation divides the sum of co-moving days by the total number of days to produce the overall percentage of days on which all four assets moved in the same direction.

Out of 1,449 total trading days, all four assets moved in the same direction on only 385 days (212 all-up + 173 all-down), or roughly 26.6% of the time. This means that on the vast majority of trading days (~73%), the four assets did not move in unison, at least one asset diverged from the others.This implies high diversification of assets within our database.

## Investigation 8: Highest-volume trading days for AMD and ETH

**Objective:** Identify the 5 highest-volume trading days for AMD and ETH individually, since volume spikes often coincide with earnings releases, major news, or sharp price moves.

```{r}
amd_top_volume <- amd_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

eth_top_volume <- eth_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

amd_top_volume
eth_top_volume
```
For each asset, a `select()` call narrows the dataset down to date, volume, and return, and `arrange()` sorts the data in descending order of volume so that the largest-volume days appear first. `head(5)` then limits each result to the top five highest-volume trading days. This process is applied separately to `amd_clean` and `eth_clean`, producing two independent tables that can be compared side by side to examine whether volume spikes for each asset tend to coincide with sharp gains or sharp losses. 

Often times, volume spikes are associated with panic selling and liquidation of stocks, but all five of AMD's highest-volume days are strongly positive, with returns ranging from +5% to a striking +23.7% on 2025-10-06. This is likely consistent with AMD's earnings-release dates or announcements/news of AI/data-center chips, thus coinciding with the boom of AI and large language models. On the other hand, ETH's highest volume days present the opposite, four out of the five returns are negative, including a severe -27.2% crash on 2021-05-19 (this aligns with the May 2021 crypto market crash) and a more recent -12.0% drop on 2025-10-10. This supports the pattern where extreme trading volume in crypto markets tends to accompany sharp sell-offs and liquidation events.

## Investigation 9: Quarterly average price levels

**Objective:** Compute the average closing price per asset by quarter to see the longer-term price trend without daily noise.

```{r}
quarterly_prices <- unified_clean |>
  mutate(year_quarter = paste0(lubridate::year(date), "-Q", lubridate::quarter(date))) |>
  group_by(year_quarter) |>
  summarise(
    avg_spy_price = mean(spy_price, na.rm = TRUE),
    avg_amd_price = mean(amd_price, na.rm = TRUE),
    avg_phi_price = mean(phi_price, na.rm = TRUE),
    avg_eth_price = mean(eth_price, na.rm = TRUE)
  )

quarterly_prices
```

A `mutate()` call creates a new `year_quarter` column by combining the year and quarter extracted from each date using `lubridate::year()` and `lubridate::quarter()`. `group_by()` then groups the data by this year-quarter label, and `summarise()` computes the average price of SPY, AMD, PHI, and ETH within each quarter using `mean()`, producing a single row per quarter that summarizes each asset's typical price level over that period. 

Across 2020-Q1 through 2024-Q4, all four assets show a clear long-term upward trend, though at different magnitudes and different volatility patterns. SPY grows steadily and relatively smoothly, from ~\$277 (2020-Q1) to ~\$577 (2024-Q4), roughly a  2.1x growth. While AMD shows much sharper swings, climbing from ~\$48 to a peak of ~\$175 (2024-Q2), but with visible mid-cycle lows, such as falling from \$133 in 2021-Q4 to \$65 by 2022-Q4. PHI has the flattest growth, moving from ~\$13 to a modest ~\$22 by 2024-Q. Lastly, ETH proves to be the most volatile, from ~\$184 (2020-Q1) to a massive peak of ~\$4,096 (2021-Q4), then crashing to ~\$1,306 by 2022-Q4, before climbing again to ~\$3,055 by 2024-Q4, roughly a 22x increase from the start of the sample.

## Investigation 10: Risk-adjusted return comparison

**Objective:** Compute a simple Sharpe ratio for each asset (average daily return minus a daily risk-free rate, divided by return standard deviation) using the 10-Year Treasury yield as the risk-free proxy, to compare returns on a risk-adjusted basis rather than raw returns alone.

```{r}
risk_free_daily <- mean(unified_clean$treasury_10y, na.rm = TRUE) / 100 / 252

risk_adjusted <- unified_clean |>
  select(amd_return, eth_return, phi_return, spy_return) |>
  summarise(
    amd_sharpe = (mean(amd_return, na.rm = TRUE) - risk_free_daily) / sd(amd_return, na.rm = TRUE),
    eth_sharpe = (mean(eth_return, na.rm = TRUE) - risk_free_daily) / sd(eth_return, na.rm = TRUE),
    phi_sharpe = (mean(phi_return, na.rm = TRUE) - risk_free_daily) / sd(phi_return, na.rm = TRUE),
    spy_sharpe = (mean(spy_return, na.rm = TRUE) - risk_free_daily) / sd(spy_return, na.rm = TRUE)
  )

risk_adjusted
```
A daily risk-free rate is first calculated outside the main query by taking the average 10-Year Treasury yield across the sample, converting it from an annual percentage to a daily decimal rate by dividing by 100 and by 252 trading days. A `select()` call then narrows the dataset down to the return columns for AMD, ETH, PHI, and SPY, and `summarise()` computes each asset's Sharpe ratio by subtracting the daily risk-free rate from its average return (`mean()`) and dividing the result by its standard deviation (`sd()`), producing a single risk-adjusted performance measure for each asset that can be compared directly regardless of differences in raw volatility. 

The Sharpe ratio measures an investment's return for each unit of risk taken, it helps in seeing if a higher return is due to smart investing or froim taking on too much risk. Even though ETH had the largest raw price appreciation over the period (see Investigation 9), SPY delivers a slightly better risk-adjusted return than ETH due to ETH's much higher volatility. AMD sits close behind both. The PHI asset, despite its steady price path in Investigation 9, actually has the weakest risk-adjusted return, its low volatility not proportional to any meaningful returns.

# Descriptive Statistics

From the previous section, a unified database named `unified_clean` was created — this is our final analytical dataset (post-join, post-fill, post-drop) that all statistics below are computed from.
```{r}
glimpse(unified_clean)
dim(unified_clean)
```

### 1. Selecting return variables
Since our end goal is an investment recommendation, we focus this analysis on the daily return series for each tradeable asset (return = risk/reward), while VIX and macro variables are analyzed separately as context in the Exploratory Financial Analysis section.
```{r}
returns_tbl <- unified_clean  |> 

  select(date,
         amd_return,
         eth_return,
         phi_return,
         spy_return)

head(returns_tbl)
dim(returns_tbl)

```

### 2. Computing median and mean
The general median and mean of returns don't give a good picture of how a "typical" return usually is due to outliers. It is just a simple average of the dataset. 
```{r}
#Mean Return
cat("AMD Mean: ")
mean(returns_tbl$amd_return,
     na.rm = TRUE)

#Median Return
cat("AMD Median: ")
median(returns_tbl$amd_return,
       na.rm = TRUE)

#Trimmed Mean
cat("AMD Trimmed Mean: ")
mean(returns_tbl$amd_return,
     trim = 0.05,
     na.rm = TRUE)

cat("ETH Mean: ")
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("ETH Median: ")
median(returns_tbl$eth_return,
       na.rm = TRUE)

cat("ETH Trimmed Mean: ")
mean(returns_tbl$eth_return,
     trim = 0.05,
     na.rm = TRUE)

cat("PHI Mean: ")
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("PHI Median: ")
median(returns_tbl$phi_return,
       na.rm = TRUE)

cat("PHI Trimmed Mean: ")
mean(returns_tbl$phi_return,
     trim = 0.05,
     na.rm = TRUE)

cat("SPY Mean: ")
mean(returns_tbl$spy_return,
     na.rm = TRUE)

cat("SPY Median: ")
median(returns_tbl$spy_return,
       na.rm = TRUE)

cat("SPY Trimmed Mean: ")
mean(returns_tbl$spy_return,
     trim = 0.05,
     na.rm = TRUE)

```

### 3. Computing Dispersion (Risk)
Calculating the variance, std dev, range, interquartile range, and coefficient of variation of returns shows a small view of each asset's general risk to return ratio.
```{r}
#Variance
cat("AMD Variance: ")
var(returns_tbl$amd_return,
    na.rm = TRUE)

#Standard Deviation
cat("AMD Standard Deviation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE)

#Range
cat("AMD Range: ")
range(returns_tbl$amd_return,
      na.rm = TRUE)

#Interquartile Range
cat("AMD IQR: ")
IQR(returns_tbl$amd_return,
    na.rm = TRUE)

#Coefficient of Variation
cat("AMD Coeff. of Variation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE) /
mean(returns_tbl$amd_return,
     na.rm = TRUE)

cat("ETH Variance: ")
var(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Standard Deviation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE)

cat("ETH Range: ")
range(returns_tbl$eth_return,
      na.rm = TRUE)

cat("ETH IQR: ")
IQR(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Coeff. of Variation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE) /
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("PHI Variance: ")
var(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Standard Deviation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE)

cat("PHI Range: ")
range(returns_tbl$phi_return,
      na.rm = TRUE)

cat("PHI IQR: ")
IQR(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Coeff. of Variation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE) /
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("SPY Variance: ")
var(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Standard Deviation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE)

cat("SPY Range: ")
range(returns_tbl$spy_return,
      na.rm = TRUE)

cat("SPY IQR: ")
IQR(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Coeff. of Variation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE) /
mean(returns_tbl$spy_return,
     na.rm = TRUE)

```

### 4. Viewing data shape
The skewness and kurtosis show the general shape of asset returns, where it is how often and how severely an asset's returns deviate from an averaged bell curve. Skewness measures the symmetry of returns, while kurtosis measures the frequency of extreme events.
```{r}
library(moments)
#Skewness
cat("AMD Skewness: ")
skewness(returns_tbl$amd_return,
         na.rm = TRUE)
#Kurtosis
cat("AMD Kurtosis: ")
kurtosis(returns_tbl$amd_return,
         na.rm = TRUE)

cat("ETH Skewness: ")
skewness(returns_tbl$eth_return,
         na.rm = TRUE)

cat("ETH Kurtosis: ")
kurtosis(returns_tbl$eth_return,
         na.rm = TRUE)

cat("PHI Skewness: ")
skewness(returns_tbl$phi_return,
         na.rm = TRUE)

cat("PHI Kurtosis: ")
kurtosis(returns_tbl$phi_return,
         na.rm = TRUE)

cat("SPY Skewness: ")
skewness(returns_tbl$spy_return,
         na.rm = TRUE)

cat("SPY Kurtosis: ")
kurtosis(returns_tbl$spy_return,
         na.rm = TRUE)
```

### 5. Computing relative data position
We calculate the minimum and maximum of each asset's return shows the biggest drop and the largest jump in value over a set time. While quartiles and percentiles of asset returns divide an asset's past returns into parts to show how often it performs well or poorly. These help in seeing an investment's risk and reward at a glance.
```{r}
#Minimum
cat("AMD Minimum: ")
min(returns_tbl$amd_return,
    na.rm = TRUE)

#Maximum
cat("AMD Maximum: ")
max(returns_tbl$amd_return,
    na.rm = TRUE)

#Quartiles
cat("AMD Quartiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

#Percentiles
cat("AMD Percentiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("ETH Minimum: ")
min(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Maximum: ")
max(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Quartiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("ETH Percentiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("PHI Minimum: ")
min(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Maximum: ")
max(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Quartiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("PHI Percentiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("SPY Minimum: ")
min(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Maximum: ")
max(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Quartiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("SPY Percentiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)
```

### 6. Dealing with Missing Values
Usually datasets have null rows that can affect the general distribuition of statistical analysis, thus null rows are to be dropped or imputed depending on the information held.
```{r}
colSums(is.na(returns_tbl))
```
Considering no null rows exist in our dataset we proceed without having to drop or impute.

### 7. Correlation Matrix
A correlation matrix of asset returns measures how investments move together and is a standardized form of the covariance that normalizes into a score between \(-1\) and \(1\).
```{r}
correlation_matrix <- returns_tbl  |> 
  select(-date)  |> 
  cor(use = "complete.obs")
correlation_matrix
```

### 8. Covariance Matrix 
A covariance matrix of asset returns also measures how investments move together,it shows the direction of the relationship between pairs of assets 
```{r}
covariance_matrix <- returns_tbl |> 
  select(-date)  |> 
  cov(use = "complete.obs")
covariance_matrix
```

### 9. Final Descriptive Statistics Table
```{r}
library(psych)

desc_stats <- psych::describe(
  returns_tbl  |> 
    select(-date)
)
round(desc_stats,4)
```

```{r}
descriptive_table <- tibble(

  Statistic = c(
    "Mean",
    "Median",
    "Trimmed Mean",
    "Standard Deviation",
    "Variance",
    "Minimum",
    "Q1",
    "Q3",
    "Maximum",
    "Range",
    "Interquartile Range",
    "Coefficient of Variation",
    "Skewness",
    "Kurtosis",
    "Observations"
  ),

  AMD = c(
    mean(returns_tbl$amd_return, na.rm = TRUE),
    median(returns_tbl$amd_return, na.rm = TRUE),
    mean(returns_tbl$amd_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE),
    var(returns_tbl$amd_return, na.rm = TRUE),
    min(returns_tbl$amd_return, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.75, na.rm = TRUE),
    max(returns_tbl$amd_return, na.rm = TRUE),
    diff(range(returns_tbl$amd_return, na.rm = TRUE)),
    IQR(returns_tbl$amd_return, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE) /
      mean(returns_tbl$amd_return, na.rm = TRUE),
    skewness(returns_tbl$amd_return, na.rm = TRUE),
    kurtosis(returns_tbl$amd_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$amd_return))
  ),
  
  ETH = c(
    mean(returns_tbl$eth_return, na.rm = TRUE),
    median(returns_tbl$eth_return, na.rm = TRUE),
    mean(returns_tbl$eth_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE),
    var(returns_tbl$eth_return, na.rm = TRUE),
    min(returns_tbl$eth_return, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.75, na.rm = TRUE),
    max(returns_tbl$eth_return, na.rm = TRUE),
    diff(range(returns_tbl$eth_return, na.rm = TRUE)),
    IQR(returns_tbl$eth_return, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE) /
      mean(returns_tbl$eth_return, na.rm = TRUE),
    skewness(returns_tbl$eth_return, na.rm = TRUE),
    kurtosis(returns_tbl$eth_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$eth_return))
  ),
  
  PHI = c(
    mean(returns_tbl$phi_return, na.rm = TRUE),
    median(returns_tbl$phi_return, na.rm = TRUE),
    mean(returns_tbl$phi_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE),
    var(returns_tbl$phi_return, na.rm = TRUE),
    min(returns_tbl$phi_return, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.75, na.rm = TRUE),
    max(returns_tbl$phi_return, na.rm = TRUE),
    diff(range(returns_tbl$phi_return, na.rm = TRUE)),
    IQR(returns_tbl$phi_return, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE) /
      mean(returns_tbl$phi_return, na.rm = TRUE),
    skewness(returns_tbl$phi_return, na.rm = TRUE),
    kurtosis(returns_tbl$phi_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$phi_return))
  ),
  
  SPY = c(
    mean(returns_tbl$spy_return, na.rm = TRUE),
    median(returns_tbl$spy_return, na.rm = TRUE),
    mean(returns_tbl$spy_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE),
    var(returns_tbl$spy_return, na.rm = TRUE),
    min(returns_tbl$spy_return, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.75, na.rm = TRUE),
    max(returns_tbl$spy_return, na.rm = TRUE),
    diff(range(returns_tbl$spy_return, na.rm = TRUE)),
    IQR(returns_tbl$spy_return, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE) /
      mean(returns_tbl$spy_return, na.rm = TRUE),
    skewness(returns_tbl$spy_return, na.rm = TRUE),
    kurtosis(returns_tbl$spy_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$spy_return))
  )
)

descriptive_table


```
# Visualization

All figures below are built from `unified_clean`, our final analytical dataset, so they stay consistent with every statistic and SQL investigation computed earlier.While numerical summaries are important in financial analysis, these are sometimes unable to capture all of the nuances of the data, particularly patterns and outliers. Visualizations thus become crucial to reveal these nuances.

## Figure 1: Indexed Price Trend (Base = 100)

Raw prices aren't directly comparable since SPY, AMD, PHI, and ETH trade on completely different price scales. Rebasing each series to start at 100 lets us compare growth trajectories side by side.
```{r}
indexed_prices <- unified_clean |>
  select(date, spy_price, amd_price, phi_price, eth_price) |>
  mutate(across(-date, ~ .x / first(.x) * 100)) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "indexed_price")

ggplot(indexed_prices, aes(x = date, y = indexed_price, color = asset)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Figure 1: Indexed Price Trend (Base = 100, Jan 2020)",
    x = "Date",
    y = "Indexed Price"
  ) +
  theme_minimal()
```

Figure 1 normalizes the assets to a base value of 100 to compare the trajectories of each asset relative to one another, across vastly different price scales. The performance is dominated by ETH, which experienced two massive macro cycles, peaking at over a 3000% return in late 2021 and again in early 2025, with a severe drawdown in 2022-2023. ETH’s extreme growth scale visually compresses the relative volatility of the traditional assets, however AMD still exhibited notable growth, peaking at a near 400% gain by late 2025. SPY and PHI both exhibited the steady, low-volatility trajectories characteristic of mature market indices and traditional equities over this six-year window.

## Figure 2: Distribution of Daily Returns by Asset

```{r}
returns_long_viz <- unified_clean |>
  select(date, amd_return, eth_return, phi_return, spy_return) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "return")

ggplot(returns_long_viz, aes(x = return, fill = asset)) +
  geom_histogram(bins = 60) +
  facet_wrap(~asset, scales = "free") +
  labs(
    title = "Figure 2: Distribution of Daily Returns by Asset",
    x = "Daily Return",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Figure 2 displays the distribution of daily returns for each asset using histograms. While all four assets have return distributions centered symmetrically around 0.0, their variances differ significantly, which visually confirms the volatility rankings from Figure 1. SPY has the narrowest range at roughly ±5%, which indicates stable, predictable daily movements. AMD and PHI show moderate variance, with wider baselines that reflect higher risk. ETH, by contrast, shows an incredibly wide distribution at ±25%, which visually demonstrates the massive macro swings observed in Figure 1.

## Figure 3: 30-Day Rolling Volatility by Asset

```{r}
rolling_vol <- unified_clean |>
  arrange(date) |>
  mutate(
    amd_vol_30d = zoo::rollapply(amd_return, width = 30, FUN = sd, fill = NA, align = "right"),
    eth_vol_30d = zoo::rollapply(eth_return, width = 30, FUN = sd, fill = NA, align = "right"),
    phi_vol_30d = zoo::rollapply(phi_return, width = 30, FUN = sd, fill = NA, align = "right"),
    spy_vol_30d = zoo::rollapply(spy_return, width = 30, FUN = sd, fill = NA, align = "right")
  ) |>
  select(date, amd_vol_30d, eth_vol_30d, phi_vol_30d, spy_vol_30d) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "rolling_vol")

ggplot(rolling_vol, aes(x = date, y = rolling_vol, color = asset)) +
  geom_line(linewidth = 0.6, na.rm = TRUE) +
  labs(
    title = "Figure 3: 30-Day Rolling Volatility by Asset",
    x = "Date",
    y = "30-Day Rolling Std. Dev. of Return"
  ) +
  theme_minimal()
```

Figure 3 gives a closer look at volatility/risk for each asset. While the previous two figures treated volatility as a constant figure, this figure shows that it is, in fact, dynamic and episodic. Previously, it was shown that ETH was consistently more volatile than the other assets. Figure 3, however, shows that its volatility instead spikes and then flatlines, becoming closer to the other assets by late 2023. This demonstrates the concept of volatility clustering, which is where volatility changes over time, and that it has a tendency to persist (Hanck et al., 2026). In other words, periods of high volatility are followed by high volatility, low volatility is followed by low volatility. This also dispels the idea that the cryptocurrency asset is always the riskiest asset in the bucket. During specific periods, like with market shifts or other corporate events, individual equities such as AMD can even become more volatile than ETH. Also of note is that all assets spiked in volatility in early 2020, likely owing to the early COVID-19 market crash. This reveals a degree of market risk that was difficult to spot in the previous figures.

## Figure 4: SPY Daily Return vs VIX Level

```{r}
ggplot(unified_clean, aes(x = vix, y = spy_return)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(
    title = "Figure 4: SPY Daily Return vs VIX Level",
    x = "VIX Level",
    y = "SPY Daily Return"
  ) +
  theme_minimal()
```

Figure 4 shifts focus from the assets to the environment around them. As mentioned earlier, SPY is a benchmark for market performance and VIX is the market “fear”. This reveals that at a low VIX level (< 20), daily returns are relatively consistent, rarely exceeding ±2%. However, as the VIX level rises, the vertical spread of the returns becomes a lot more volatile.

## Figure 5: Correlation Heatmap of Daily Returns

```{r}
corr_df <- as.data.frame(correlation_matrix) |>
  tibble::rownames_to_column("asset1") |>
  tidyr::pivot_longer(-asset1, names_to = "asset2", values_to = "correlation")

ggplot(corr_df, aes(x = asset1, y = asset2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", midpoint = 0) +
  labs(
    title = "Figure 5: Correlation Heatmap of Daily Returns",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()
```

Figure 5 evaluates the co-movement of assets and potential for diversification. All pairwise correlations are positive, ranging from weak (r=0.16) to moderate-high (r=0.64). The strongest relationship is between AMD and SPY, which reflects high systematic risk exposure for AMD. PHI displays consistently low correlations with the other assets, which makes sense as it is Philippine-based. This exposes it to different risks when compared to the other assets.

## Figure 6: Cumulative Growth of $1 Invested

```{r}
cumulative_growth <- unified_clean |>
  arrange(date) |>
  mutate(
    amd_cum = cumprod(1 + coalesce(amd_return, 0)),
    eth_cum = cumprod(1 + coalesce(eth_return, 0)),
    phi_cum = cumprod(1 + coalesce(phi_return, 0)),
    spy_cum = cumprod(1 + coalesce(spy_return, 0))
  ) |>
  select(date, amd_cum, eth_cum, phi_cum, spy_cum) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "growth_of_1")

ggplot(cumulative_growth, aes(x = date, y = growth_of_1, color = asset)) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  labs(
    title = "Figure 6: Cumulative Growth of $1 Invested (2020-2025)",
    x = "Date",
    y = "Value of $1 Invested"
  ) +
  theme_minimal()
```

Figure 6 focuses on wealth outcomes from investments in a portfolio. ETH reaches large peaks in late 2021 at nearly $12, but settles at around $5 by the end of 2025. By contrast, AMD steadily grew to around that, without the massive drawdowns in between. This risk of drawdown highlights how difficult it is to capture that wealth without active risk management. Additionally, PHI notably lagged behind the SPY benchmark, which shows how an asset can underperform these broad market benchmarks over a long time horizon.

## Figure 7: Daily Return Distribution by Asset (Boxplot)

```{r}
ggplot(returns_long_viz, aes(x = asset, y = return, fill = asset)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Figure 7: Daily Return Distribution by Asset",
    x = "Asset",
    y = "Daily Return"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Figure 7 displays all the return distributions on the same y-axis scale. AMD and ETH notably have taller central boxes, which represent the middle 50% of daily returns. Notably, this visualization helps to visualize the outliers to expose each asset’s value at risk (VaR). ETH shows an extreme outlier data point stretching down to just over a 40% loss. Additionally, this shows a hidden risk to the PHI asset, which up to now has displayed generally conservative, safe characteristics. When compared to SPY, PHI displays negative asymmetry, which indicates that extreme drawdown days impact PHI more than SPY.

## Figure 8: SPY Price vs CPI and Fed Funds Rate Over Time

```{r}
macro_long <- unified_clean |>
  select(date, spy_price, cpi, fedfunds) |>
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value")

ggplot(macro_long, aes(x = date, y = value)) +
  geom_line(color = "darkslategray", na.rm = TRUE) +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(
    title = "Figure 8: SPY Price vs CPI and Fed Funds Rate Over Time",
    x = "Date",
    y = NULL
  ) +
  theme_minimal()
```

Figure 8 once again focuses on the macroeconomic environment. This aligns SPY performance with inflation, measured by CPI, and monetary policy, measured by the Federal Funds rate. This chart captures the aggressive policy change in early 2022 where the short-term interest rates were raised to over 5% in response to the sharp rise in CPI post-2020.

# Exploratory Financial Analysis

This section does not recompute new statistics from scratch. It pulls together what was already built in the SQL Investigation and Visualization sections and interprets them together, since the spec asks us to compare across five specific angles: returns, volatility, inflationary periods, high-VIX periods, and interest-rate environments.

## 1. Comparing Returns and Volatility Across Assets

```{r}
return_vol_summary <- returns_long_viz |>
  group_by(asset) |>
  summarise(
    avg_daily_return = mean(return, na.rm = TRUE),
    annualized_return = mean(return, na.rm = TRUE) * 252,
    daily_volatility = sd(return, na.rm = TRUE),
    annualized_volatility = sd(return, na.rm = TRUE) * sqrt(252)
  ) |>
  arrange(desc(annualized_return))

return_vol_summary
```

Annualizing the daily figures for 252 trading days puts all four assets on a comparable yearly basis instead of the tiny daily numbers that are hard to reason about directly. Pairing this table with Figure 1 (indexed price trend) and Figure 6 (cumulative growth of $1), an asset can have a strong annualized return on paper while Figure 6 shows the path there was rough with deep drawdowns along the way. Figure 7's boxplot is the fastest way to see this same return-vs-volatility tradeoff visually, a tall box means more day-to-day risk for whatever return the asset ultimately delivered.



## 2. Inflationary Periods

```{r}
cpi_trend
```

**Interpretation:** This reuses the CPI-rising vs CPI-flat/falling split from SQL Investigation 6. Since CPI only updates monthly and we forward-filled it across each trading day, "CPI Rising" here means the whole month following a higher print, not a single release-day reaction. Cross-reference this with Figure 8, where the CPI panel shows the actual inflation path over 2020-2025, to see whether the rising-CPI periods in the table line up with the sharpest sections of that line. If equity and crypto returns are consistently weaker during CPI Rising months, that is direct evidence inflation surprises pressure risk assets in this dataset, not just a theoretical assumption.

## 3. High-VIX Periods

```{r}
vix_regime_stats
```

**Interpretation:** This reuses the High VIX vs Low VIX split from SQL Investigation 4, built around the median VIX level in our sample. Figure 3 (30-day rolling volatility) and Figure 4 (SPY return vs VIX scatter) are the visual backbone for this comparison, Figure 3 shows exactly when the market moved into a High VIX regime, and Figure 4 shows whether higher VIX levels actually coincided with weaker or more unpredictable SPY returns on a day-by-day basis rather than just in aggregate. If both the standard deviation and the average return move in the expected direction (higher volatility, lower average return) during High VIX days, that is a strong basis for treating VIX as an early warning signal in the investment recommendation, not just a descriptive curiosity.

## 4. Interest-Rate Environments

```{r}
fedfunds_trend
```

**Interpretation:** This reuses the Rate Hike vs Rate Cut vs Held Steady split from SQL Investigation 5. Crypto is generally the most sensitive of the four assets to rate changes, since it carries no yield of its own to compete with a higher risk-free rate, so ETH's average return during Rate Hike days is worth checking closely against AMD, PHI, and SPY. Figure 8's Fed Funds panel shows the actual rate path across the sample, use it to see whether the current environment (as of the end of the sample period) sits in a hiking, cutting, or holding phase, since that context should carry directly into how much crypto exposure the investment recommendation allocates going forward.

## 5. Bringing the Comparisons Together

**Interpretation:** Across all four lenses, average and annualized returns (Section 1), inflationary periods (Section 2), high-VIX periods (Section 3), and interest-rate environments (Section 4), the goal is to identify which asset holds up best under stress (High VIX, CPI Rising, Rate Hike) rather than only which asset has the best raw average return in calm conditions. An asset that leads on annualized return but falls apart hardest under all three stress conditions is a very different allocation decision than one that trails slightly on raw return but stays comparatively stable. This is the evidence base the Investment Recommendation section should draw from directly, rather than restating the descriptive statistics table in isolation.

# Investment Recommendation

We suggest a 60/30/10 split of SPY, AMD, and ETH-USD respectively with a total investment fund of USD 1,000,000. A moderate risk appetite would act as the basis for this recommendation, where the aim is to split the allocated funds in a practical way, backed by analytical evidence of investigated data.  Considering each asset,

### S&P 500 (SPY) - Allocation: USD 600,000 (60%)
SPY acts as the foundation for this portfolio, as it already contains structural diversification with consistent risk-adjusted performance. Across the sample period, SPY posted the highest Sharpe ratio of all four assets (0.0471), meaning it delivered the best return per unit of risk taken even with ETH’s larger raw price appreciation. SPY's volatility was also consistently the lowest across regimes when compared against other assets through analysis of VIX regimes and standard deviation. Given this combination of the strongest risk-adjusted return and the lowest volatility profile in the dataset, SPY is allocated the majority of the portfolio to establish a stable baseline.

### Advanced Micro Devices, Inc. (AMD) - Allocation: USD 300,000 (30%)
AMD was allocated the second largest share of the portfolio because it offers higher return potential than SPY, while still exhibiting return patterns broadly aligned with the broader market. AMD's Sharpe ratio (0.0423) was close to SPY's, indicating it has historically rewarded its additional volatility reasonably well rather than adding risk without compensated returns, but AMD’s higher beta was evident throughout the analysis. Because it is relatively more correlated to SPY, when SPY experiences movement or downturns AMD has more dramatic swings due to its higher volatility. However, AMD also experiences sector-specific upturns, especially with the boom of AI and its needed hardware specifications, making AMD a suitable secondary holding with meaningful and expected growth potential.

### Ethereum (ETH-USD) - Allocation: USD 100,000 (10%)
ETH was given the smallest allocation due to its more extreme risk profile out of all the assets in the database. ETH’s Sharpe ratio (0.0456) is close to both SPY and AMD, indicating it has still historically compensated reasonably for its risk under the 5 year period. Despite the relatively “stable” Sharpe ratio, looking at ETH’s price history, it is clearly highly volatile, with massive growths and massive crashes creating a high risk profile. An allocation of 10% is intended to capture ETH’s upside and low correlation with SPY and AMD on most days, it is kept small enough so extreme crashes will not severely affect the health of the overall portfolio.

### PLDT Inc. ADR (PHI) - Allocation: USD 0 (0%)
PHI was excluded from the final allocation due to its comparatively weak risk-adjusted return, it posted the lowest Sharpe ratio (0.0198) among all four assets, with quarterly price data showing long stretches of stagnated movement. Although PHI does consistently show independence from U.S. market stress, going off a moderate risk appetite, we prioritized assets with demonstrated strength in their risk-adjusted over pure diversification benefits.

### Cash - Allocation: USD 0 (0%)
No portion of the portfolio was held in cash, as the objective of our analysis paired with our risk appetite, was to fully deploy the available USD 1,000,000 into assets with positive expected returns rather than hold idle capital. While cash offers stability and eliminates market-related volatility, it also guarantees a loss of real purchasing power over time due to inflation or other macroeconomic factors.

This 60/30/10 split concentrates capital in the highest risk-adjusted performer (SPY), we use AMD to add market-correlated growth exposure, but with amplified risk. ETH is added to create exposure to benefit from its high upside and low correlation on most days while containing its tail risk. This results in a portfolio designed to outperform a passive single asset portfolio by balancing needed diversification and risk.


# References
- Board of Governors of the Federal Reserve System (US). Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity, Quoted on an Investment Basis [DGS10]. Federal Reserve Bank of St. Louis. (2026).  https://fred.stlouisfed.org/series/DGS10
- Chen, J. (2026). Price Discovery Explained: Process, Factors, and Comparison with Valuation. https://www.investopedia.com/terms/p/pricediscovery.asp
- Hanck, C., Arnold, M., Gerber, A., Schmelzer, M. (2026). Volatility Clustering and Autoregressive Conditional Heteroskedasticity | Introduction to Econometrics in R. Introduction to Econometrics in R. https://www.econometrics-with-r.org/16.4-volatility-clustering-and-autoregressive-conditional-heteroskedasticity.html
- Hayes, A. (2025). Financial Markets: Role in the Economy, Importance, Types, and Examples. https://www.investopedia.com/terms/f/financial-market.asp
- Semiconductor Industry Association. (2026). Industry Impact - Semiconductor Industry Association. https://www.semiconductors.org/industry-impact/
- U.S. Bureau of Labor Statistics. (2026). Consumer Price Index for All Urban Consumers: All Items in U.S. City Average [CPIAUCSL]. Federal Reserve - - Bank of St. Louis. https://fred.stlouisfed.org/series/CPIAUCSL
- Yahoo Finance. (2026). PLDT Inc. (PHI) Stock Price, News, Quote & History - Yahoo Finance. https://finance.yahoo.com/quote/PHI/
- Hayes, A. (2026, January 23). Understanding Liquidity and How to Measure It. Investopedia. https://www.investopedia.com/terms/l/liquidity.asp
- Gopalakrishnan, J. (2025, December 10). Stock investment risk: Understanding systematic and idiosyncratic risks. Encyclopedia Britannica. https://www.britannica.com/money/types-of-investment-risks
- Gratton, P. (2026, April 27). What Is the Stock Market and How Does It Work? Investopedia. https://www.investopedia.com/terms/s/stockmarket.asp#toc-what-is-a-stock-exchange
- O’Trakoun, J. (2025, December 15). Phantom figures: missing data in October. Richmond Fed. https://www.richmondfed.org/research/national_economy/macro_minute/2025/phantom_figures_missing_data_in_october
