---
title: "FINSTAD Financial Database"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk, zoo)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
library(zoo)
```

### Comparing the different datasets

We first get symbols from the following sources:
- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis' FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("FEDFUNDS",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

Then we can look at the macroeconomic data,

The data is taken from the Federal Reserve Economic Database (FRED) from the Federal Reserve Bank of St. Louis, where we load in the _Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity_ (**DGS10**), _Consumer Price Index for All Urban Consumers_ (**CPIAUCSL**), and the _Federal Funds Effective Rate_ (**FEDFUNDS**).

**DGS10** is used as a proxy for the "risk-free rate" in finance, it helps investors gauge inflation expectations, economic growth, and monetary policy shifts. **CPIAUCSL** is seasonally adjusted to gauge monthly inflation metrics on consumer goods. **FEDFUNDS** is the monthly average interest rate at which banks lend reserves to each other overnight, and reflects the stance of U.S. monetary policy (tightening vs. easing cycles).

These pieces of macroeconomic data would help in understanding broader market cycles and how systemic risks affect stock valuations.
```{r}
head(DGS10)
head(CPIAUCSL)
head(FEDFUNDS)
```

We can then take a look at our chosen financial market data,

U.S. Asset - Advanced Micro Devices, Inc. (AMD)
Philippine Asset - PLDT, Inc.(PHI)
Cryptocurrency - Etherium (ETH or ETH-USD)
Required - State Street SPDR S&P 500 ETF Trust (SPY)

```{r}
ETH <- `ETH-USD`

head(AMD)
head(PHI)
head(ETH)
head(SPY)
```

Lastly, we take a look at the _Cboe Volatility Index_ (**VIX**), this is not an individual stock but a real time market index created by the Chicago Board Options Exchange (Cboe). This measures the market's expectation of 30-day volatility in the S&P 500 index.

This adds aid on top of macroeconomic data to further gauge systemic risks, VIX is known as the "Fear Gauge" which reflects levels of anxiety, complacency, or uncertainty in the market, when this "Fear Gauge" level rises it implies larger expected swings in stock prices.
```{r}
head(VIX)
```

# Cleaning

### 1. Converting XTS to tibbles
First we convert all XTS objects to tibbles for easier data manipulation.
```{r}
dgs10_tbl <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl_tbl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy_tbl <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
  adj_close = as.numeric(Ad(SPY)),
)
amd_tbl <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
  adj_close = as.numeric(Ad(AMD)),
)
phi_tbl <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
  adj_close = as.numeric(Ad(PHI)),
)
vix_tbl <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
  adj_close = as.numeric(Ad(VIX)),
)
eth_tbl <- tibble(
  date = index(ETH),
  open = as.numeric(ETH$`ETH-USD.Open`),
  high = as.numeric(ETH$`ETH-USD.High`),
  low = as.numeric(ETH$`ETH-USD.Low`),
  close = as.numeric(ETH$`ETH-USD.Close`),
  volume = as.numeric(ETH$`ETH-USD.Volume`),
  adjusted = as.numeric(ETH$`ETH-USD.Adjusted`),
  adj_close = as.numeric(Ad(ETH)),
)
```

### 2. Asset Identifier
We add an asset identifier per asset to see the data better in table structures.
```{r}
amd_tbl <- amd_tbl |> 
  mutate(asset = "AMD")

eth_tbl <- eth_tbl |> 
  mutate(asset = "ETH")

phi_tbl <- phi_tbl |> 
  mutate(asset = "PHI")

spy_tbl <- spy_tbl |> 
  mutate(asset = "SPY")

vix_tbl <- vix_tbl |> 
  mutate(asset = "VIX")
```

### 3. Computing daily returns

Daily returns are computed using the basic return formula
```{r}
amd_tbl <- amd_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

eth_tbl <- eth_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

phi_tbl <- phi_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

spy_tbl <- spy_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

vix_tbl <- vix_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )
```

### 4. Removing missing rows
Null rows are dropped to preserve data integrity
```{r}
amd_tbl <- amd_tbl  |> 
  filter(!is.na(return))

eth_tbl <- eth_tbl  |> 
  filter(!is.na(return))

phi_tbl <- phi_tbl |> 
  filter(!is.na(return))

spy_tbl <- spy_tbl  |> 
  filter(!is.na(return))

vix_tbl <- vix_tbl  |> 
  filter(!is.na(return))
```

### 5. Examine the data
```{r}
glimpse(amd_tbl)
glimpse(eth_tbl)
glimpse(phi_tbl)
glimpse(spy_tbl)
glimpse(vix_tbl)
```

Summary Tables
```{r}
summary(amd_tbl)
summary(eth_tbl)
summary(phi_tbl)
summary(spy_tbl)
summary(vix_tbl)
```

### 6. Preparing Macroeconomic variables
```{r}
treasury_tbl <- tibble(
  date = index(DGS10),
  treasury_10y = as.numeric(DGS10$DGS10)
)
```

```{r}
cpi_tbl <- tibble(
  date = index(CPIAUCSL),
  cpi = as.numeric(CPIAUCSL$CPIAUCSL)
)
```

```{r}
fedfunds_tbl <- tibble(
  date = index(FEDFUNDS),
  fedfunds = as.numeric(FEDFUNDS$FEDFUNDS)
)
```

```{r}
head(treasury_tbl)
head(cpi_tbl)
head(fedfunds_tbl)
```

### 7. Create an Observation Summary Table and saving it
We first count the observations
```{r}
nrow(amd_tbl)
nrow(eth_tbl)
nrow(phi_tbl)
nrow(spy_tbl)
nrow(vix_tbl)
nrow(treasury_tbl)
nrow(cpi_tbl)
nrow(fedfunds_tbl)
```

Creating the observation summary table
```{r}
obs_summary <- tibble(
  Dataset = c("AMD",
              "Ethereum",
              "PLDT",
              "SPY",
              "VIX",
              "Treasury",
              "CPI",
              "Fed Funds Rate"),
  Observations = c(
    nrow(amd_tbl),
    nrow(eth_tbl),
    nrow(phi_tbl),
    nrow(spy_tbl),
    nrow(vix_tbl),
    nrow(treasury_tbl),
    nrow(cpi_tbl),
    nrow(fedfunds_tbl)
  )
)

obs_summary
```

Saving the clean tables
```{r}
amd_clean <- amd_tbl
eth_clean <- eth_tbl
phi_clean <- phi_tbl
spy_clean <- spy_tbl
vix_clean <- vix_tbl
treasury_clean <- treasury_tbl
cpi_clean <- cpi_tbl
fedfunds_clean <- fedfunds_tbl
```

# Database Integration

After cleaning all the data, we collate the rows into 1 unified dataset through the usage of SQL joins. We first need to prepare the necessary tables for unifying the datasets.

## 1. Renaming variables
```{r}
amd_sql <- amd_clean  |> 
  select(date,
         amd_price = adj_close,
         amd_return = return)

eth_sql <- eth_clean  |> 
  select(date,
         eth_price = adj_close,
         eth_return = return)

phi_sql <- phi_clean  |> 
  select(date,
         phi_price = adj_close,
         phi_return = return)

spy_sql <- spy_clean  |> 
  select(date,
         spy_price = adj_close,
         spy_return = return)

vix_sql <- vix_clean |> 
  select(date,
         vix = adj_close,
         vix_return = return)
```

```{r}
head(amd_sql)
head(eth_sql)
head(phi_sql)
head(spy_sql)
head(vix_sql)
```

## 2. The Trading Calendar Problem

Our four core assets do not trade on the same schedule:

- **SPY** and **AMD** follow the NYSE/NASDAQ calendar (U.S. weekdays, minus U.S. market holidays).
- **PHI** (PLDT ADR) also trades on a U.S. exchange (NYSE), so it largely follows the same calendar as SPY/AMD, but can still diverge on days one market is closed and the other isn't (e.g. differing holiday schedules).
- **ETH** trades every single calendar day, including weekends and all holidays, since crypto markets never close.
- **VIX** follows the CBOE/NYSE calendar, similar to SPY.

This means a naive join on `date` will not line up perfectly — ETH alone contributes roughly 30% more calendar dates per year than the equity assets. This mismatch is exactly why the join strategy matters, and why we need to demonstrate multiple join types before picking one.

## 3. Demonstrating Join Types

Using `spy_sql` (equity calendar) and `eth_sql` (7-day calendar) as the clearest illustration of the mismatch:

```{r}
cat("SPY rows: ", nrow(spy_sql), "\n")
cat("ETH rows: ", nrow(eth_sql), "\n")
```

**`inner_join()`** — keeps only dates present in *both* datasets. This throws away every weekend/holiday ETH observation that SPY doesn't have.
```{r}
demo_inner <- spy_sql |> 
  inner_join(eth_sql, by = "date")

cat("inner_join rows: ", nrow(demo_inner), "\n")
```

**`full_join()`** — keeps every date present in *either* dataset, filling the missing side with `NA`. This preserves ETH's weekend rows but introduces `NA` for SPY/AMD/PHI on those same days, since equities simply didn't trade.
```{r}
demo_full <- spy_sql |> 
  full_join(eth_sql, by = "date")

cat("full_join rows: ", nrow(demo_full), "\n")
cat("Rows where SPY is NA (i.e. weekend/holiday-only ETH dates): ",
    sum(is.na(demo_full$spy_price)), "\n")
```

**`anti_join()`** — shows *only* the rows that don't have a match, which is useful for auditing exactly which dates get dropped or orphaned by a join. Here, it shows every date ETH traded that SPY didn't:
```{r}
demo_anti <- eth_sql |> 
  anti_join(spy_sql, by = "date")

cat("Dates ETH traded but SPY did not: ", nrow(demo_anti), "\n")
head(demo_anti)
```

**`left_join()`** — anchors on the left-hand table's dates and pulls in matches from the right, discarding right-only rows. Anchoring on `spy_sql` keeps every SPY trading day and attaches ETH's price on that date where available:
```{r}
demo_left <- spy_sql |> 
  left_join(eth_sql, by = "date")

cat("left_join rows: ", nrow(demo_left), "\n")
cat("Rows where ETH is NA: ", sum(is.na(demo_left$eth_price)), "\n")
```

## 4. Final Join Strategy

Our final database is meant to answer an **equity/benchmark-driven investment question** — comparing AMD, PHI, and ETH against SPY as the reference benchmark, alongside macro conditions. Given that framing:

- `inner_join()` is too aggressive — it would silently drop every day where even one asset didn't trade, shrinking the dataset unnecessarily and losing information about days SPY *did* move.
- `full_join()` keeps the most data, but introduces weekend/holiday rows where SPY, AMD, PHI, and VIX are all structurally `NA` (they simply have no observation those days) — this isn't "missing" data to be filled, it's a non-event, so those extra rows have no analytical value for an equity-anchored study.
- `anti_join()` is a diagnostic tool, not a database-building join — we use it above to confirm exactly how many ETH-only dates exist, but it doesn't belong in the final chain.
- **`left_join()` anchored on `spy_sql`** is the best choice, it preserves every SPY trading day (our benchmark and reference calendar), attaches whichever other series have data on that date, and cleanly produces `NA` only where a genuine data gap exists (e.g. ETH trading on a day AMD didn't, or a macro series not yet updated) rather than manufacturing rows that don't correspond to any equity trading activity.

## 5. Building the Unified Database
```{r}
unified_db <- spy_sql |> 
  left_join(amd_sql,
            by = "date") |> 
  left_join(phi_sql,
            by = "date") |>
  left_join(eth_sql,
            by = "date") |> 
  left_join(vix_sql,
            by = "date") |> 
  left_join(treasury_clean,
            by = "date") |> 
  left_join(cpi_clean,
            by = "date") |> 
  left_join(fedfunds_clean,
            by = "date") |> 
  arrange(date) |>
  tidyr::fill(treasury_10y,
              cpi,
              fedfunds,
              .direction = "downup")

glimpse(unified_db)
head(unified_db)
view(unified_db)
```

`treasury_10y` and `fedfunds` are only reported on U.S. business days, while `cpi` is only reported once a month. Since our final database is anchored on SPY's daily trading calendar, joining these lower-frequency series directly leaves `NA` on every date. We resolve this with `tidyr::fill(..., .direction = "downup")`, which carries the most recent known value forward through time (e.g. a CPI print from early June applies to every trading day until the next print in early July), and also back fills any leading rows at the very start of the sample that come before the first available reading. Using only `"down"` would leave those first few weeks of 2020 as `NA`, since there is nothing earlier to carry forward from. This is standard practice for macro series in daily financial datasets, since the "current" CPI or Fed Funds level doesn't change on the days between official releases, and a short leading gap is filled with the nearest known value rather than left blank.

## 6. Observation Tracking

Tracking row counts at each stage of the join chain shows exactly where (and why) the dataset shrinks or holds steady.
```{r}
obs_track_spy    <- spy_sql
obs_track_amd    <- obs_track_spy |> left_join(amd_sql, by = "date")
obs_track_phi    <- obs_track_amd |> left_join(phi_sql, by = "date")
obs_track_eth    <- obs_track_phi |> left_join(eth_sql, by = "date")
obs_track_vix    <- obs_track_eth |> left_join(vix_sql, by = "date")
obs_track_treas  <- obs_track_vix |> left_join(treasury_clean, by = "date")
obs_track_cpi    <- obs_track_treas |> left_join(cpi_clean, by = "date")
obs_track_ff     <- obs_track_cpi |> left_join(fedfunds_clean, by = "date")

join_tracking <- tibble(
  Step = c("Original SPY (anchor)",
           "+ AMD",
           "+ PHI",
           "+ ETH",
           "+ VIX",
           "+ Treasury (DGS10)",
           "+ CPI",
           "+ Fed Funds Rate"),
  Rows_After_Join = c(
    nrow(obs_track_spy),
    nrow(obs_track_amd),
    nrow(obs_track_phi),
    nrow(obs_track_eth),
    nrow(obs_track_vix),
    nrow(obs_track_treas),
    nrow(obs_track_cpi),
    nrow(obs_track_ff)
  )
)

join_tracking
```

Because every join in the chain is a `left_join()` anchored on `spy_sql`, **row count never shrinks past the original SPY row count** — a `left_join()` can only add columns/matches or introduce `NA`, never drop rows from the left-hand side. This is exactly the guarantee we wanted from our join strategy: the equity benchmark's trading calendar is preserved intact all the way through.

Checking null rows after the join and fill step:
```{r}
colSums(
  is.na(
    unified_db
  )
)
```

## 7. Dropping Null Rows

After `tidyr::fill(..., .direction = "downup")`, the macro columns (`treasury_10y`, `cpi`, `fedfunds`) should have zero remaining `NA`, including at the very start of the sample. Any `NA` still present at this point belongs to the **asset price/return columns** (`amd_price`, `amd_return`, `phi_price`, `phi_return`, `eth_price`, `eth_return`, `vix`, `vix_return`) — these represent real gaps where a given asset simply didn't trade on a day SPY did (e.g. a PSE holiday that isn't a NYSE holiday), not something we can reasonably forward-fill without fabricating price movement that never happened.

We drop these remaining rows, since imputing a stock/crypto price would distort the return and volatility statistics that our entire analysis depends on:
```{r}
unified_clean <- unified_db |> 
  tidyr::drop_na(amd_price, amd_return,
                 phi_price, phi_return,
                 eth_price, eth_return,
                 vix, vix_return,
                 spy_price, spy_return)

cat("Rows before dropping asset-level NA: ", nrow(unified_db), "\n")
cat("Rows after dropping asset-level NA: ", nrow(unified_clean), "\n")
cat("Rows dropped: ", nrow(unified_db) - nrow(unified_clean), "\n")

colSums(is.na(unified_clean))
```

`unified_clean` is our final analytical dataset, used for all descriptive statistics, SQL investigations, and visualizations going forward.

# SQL Investigation

Using the SQL-based query functions select(), filter(), mutate(), arrange(), group_by() and summarise(), we investigate 10 financial questions to broaden understanding of observed relationships within the dataset.

## Investigation 1: Biggest single-day moves per asset

**Objective:** Find the top 3 single-day returns for each of the four core assets to see which asset produces the largest short-term stocks.
```{r}
returns_long <- unified_clean |>
  select(date, amd_return, eth_return, phi_return, spy_return) |>
  tidyr::pivot_longer(cols = -date, names_to = "asset", values_to = "return")

top_gains <- returns_long |>
  group_by(asset) |>
  arrange(desc(return), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  filter(rank <= 3) |>
  select(asset, date, return, rank)

top_gains
```

Using the new unified dataset, we want to see which 3 days within the time period provided the highest returns per asset. By first using select(), only the necessary columns are selected (date, amd_return, eth_return, phi_return, spy_return) where it is then grouped by asset name and arranged/sorted by highest returns. The mutate() function then assigns a new ranking column and each asset is filtered to only the top 3.

As seen in the resulting table, AMD and ETH-USD are the 2 assets with the highest returns, averaging almost double the PHI asset. The returns are relatively similar, indicating a regular historical period with no outliers during this specific time period.

## Investigation 2: Monthly average returns

**Objective:** Compute the average daily return per calendar month for each asset to identify which months or periods were consistently strong or weak.
```{r}
monthly_avg_returns <- unified_clean |>
  mutate(year_month = format(date, "%Y-%m")) |>
  group_by(year_month) |>
  summarise(
    avg_amd = mean(amd_return, na.rm = TRUE),
    avg_eth = mean(eth_return, na.rm = TRUE),
    avg_phi = mean(phi_return, na.rm = TRUE),
    avg_spy = mean(spy_return, na.rm = TRUE)
  )

monthly_avg_returns
```

Using the new unified dataset, we compute an average of the daily return to see which months are strong or weakpoints for each asset in our database. Using the mutate() function, the dates are formatted from YYYY-MM-DD to YYYY-MM to prepare for the grouping of data by months and subsequent averaging of daily returns. Due to the amount of rows, we can simplify the viewing of the lowest and highest months per year per asset where we use group_by() per asset and return the best and worst performing months.

```{r}
monthly_long <- monthly_avg_returns |>
  tidyr::pivot_longer(
    cols = c(avg_amd, avg_eth, avg_phi, avg_spy),
    names_to = "asset",
    values_to = "avg_return"
  )

best_months <- monthly_long |>
  group_by(asset) |>
  slice_max(avg_return, n = 1) |>
  rename(best_month = year_month, best_return = avg_return)

worst_months <- monthly_long |>
  group_by(asset) |>
  slice_min(avg_return, n = 1) |>
  rename(worst_month = year_month, worst_return = avg_return)

best_worst_months <- best_months |>
  left_join(worst_months, by = "asset") |>
  select(asset, best_month, best_return, worst_month, worst_return)

best_worst_months
```

Although the best and worst months of average returns within a time period is not the best indicator of general asset behavior, we can view it through a historical financial lense. The best months have little to no correlation with each other, but there is a clear pattern in the worst year and month of SPY, ETH, and AMD This shared bad performance could be due to the 2022 stock market decline (https://en.wikipedia.org/wiki/2022_stock_market_decline) that affected several stock market indices worldwide between January and October 2022. Many major world events occurred during this period, such as the Russian invasion of Ukraine and the effects of the Covid-19 pandemic on the economy. These caused inflation readings to rise with general fear of an oncomign recession. 

## Investigation 3: SPY selloff days and cross-asset behavior

**Objective:** Identify SPY's 10 worst trading days and check how AMD, PHI, and ETH performed on those same dates, to test whether any of them behave as a diversifier during U.S. market selloffs.
```{r}
spy_selloffs <- unified_clean |>
  filter(spy_return < 0) |>
  arrange(spy_return) |>
  select(date, spy_return, amd_return, phi_return, eth_return) |>
  head(10)

spy_selloffs
```

Taking a closer look at specific dates over daily averages over the period of a month could provide more insight towards the correlation of asset movement and SPY movement. Using the unified database, SPY returns are filtered to negative dates and arranged/sorted by lowest returns. Asset return columns are added for easier comparison of correlation between SPY's worst days and asset performance on that day. 

Taking a look at the returns on SPY's worst days, there seems to be a correlation with negative returns in most asset dates, but it does not seem to be _highly_ correlated, as each assets' worst return within the calendar snapshot does not necessarily align with SPY's ranked worst days. Relative to the other returns, AMD seems to be the most correlated with SPY which makes sense because it is part of the S&P 500. Low to moderate correlation is good, as it means the asset portfolio is quite diversified.

## Investigation 4: High-VIX vs low-VIX return and volatility comparison

**Objective:** Split trading days into High VIX and Low VIX regimes (relative to the median VIX level) and compare average return and volatility in each regime.

```{r}
vix_median <- median(unified_clean$vix, na.rm = TRUE)

vix_regime_stats <- unified_clean |>
  mutate(vix_regime = if_else(vix > vix_median, "High VIX", "Low VIX")) |>
  group_by(vix_regime) |>
  summarise(
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    sd_spy_return = sd(spy_return, na.rm = TRUE),
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    sd_amd_return = sd(amd_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    sd_phi_return  = sd(phi_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    sd_eth_return  = sd(eth_return, na.rm = TRUE),
    n_days = n()
  )

vix_regime_stats
```
Considering VIX calculates market volatility, we take the average and the standard deviation of returns for assets and SPY to assess average daily returns and the magnitude of swings. We first separate the data into high and low VIX periods using mutate(), the values in those periods are the aggregated asset data grouped by VIX regimes and summarized based on statistical computations.

As seen in the resulting table the high and low VIX instances are relatively equal, making this a fair assessment on how VIX is able to act as a "fear gauge". 'High VIX' consistently has a lower average of returns and low VIX being the opposite, with the assets standard deviation also following the same pattern, larger volatility/swing in high VIX and lower in low VIX periods. This shows how VIX is able to accurately gauge market volatility.

## Investigation 5: Returns across Fed Funds rate regimes

**Objective:** Classifying each trading day by whether the Fed Funds Rate was rising, falling, or holding steady, then compare average returns across the three regimes to gauge rate sensitivity.

```{r}
fedfunds_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    fedfunds_change = fedfunds - lag(fedfunds),
    rate_regime = case_when(
      fedfunds_change > 0 ~ "Rate Hike",
      fedfunds_change < 0 ~ "Rate Cut",
      TRUE ~ "Held Steady"
    )
  ) |>
  filter(!is.na(rate_regime)) |>
  group_by(rate_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

fedfunds_trend
```

To see the relationship between our assets and FEDFUNDS, we compare average returns to the status of FEDFUNDS based on their change column. Using the unified database, the change column is assigned a "rate class" to create groups for data to be aggregated by, the average returns are computed using summarize() then grouped by the rate class produced earlier. 

In the resulting table, there is a clear imbalance between the number of days per rate class but this is expected as FEDFUNDS is a policy rate, not a continuously trading price, hence majority of days will naturally fall under the "Held Steady" class because rate changes are rare, discrete events.This also makes average returns sensitive to rate cuts and hikes, despite the small sample sizes for Rate Hike and Rate Cut days, a consistent pattern shows where every asset earns its highest average return during Rate Cut periods, and its lowest/near-lowest average during "Held Steady" periods. However, this pattern must be interpreted cautiously, as with only 8 Rate Cut days and 17 Rate Hike days in the entire sample, these averages are highly sensitive to one or two extreme trading days rather than reflecting a strong statistical relation.

## Investigation 6: Returns across inflation regimes

**Objective:** Classify each trading day by whether CPI was rising or falling/flat that month, then compare average returns across the two inflation regimes.

```{r}
cpi_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    cpi_change = cpi - lag(cpi),
    inflation_regime = if_else(cpi_change > 0, "CPI Rising", "CPI Falling or Flat")
  ) |>
  filter(!is.na(inflation_regime)) |>
  group_by(inflation_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

cpi_trend
```

Now looking at the relationship between our assets and CPIAUCSL, we can check whether inflation directly affects the returns of stocks. Using the unified database, CPI changes are grouped based on change column values and sorted into classes.Using the group_by() function, we aggregate the return data of our assets through summarise() into the mentioned classes.

Looking at the resulting table, the number of days per class remains imbalanced, but this is expected as CPI is a monthly economic indicato and forward-filled onto the SPY trading calendar. Despite the small sample for "CPI Rising" a consistent pattern emerges with SPY, ETH, and PHI, where they all show noticeably higher average returns during CPI Rising periods compared to CPI Falling or Flat. AMD on the other hand shows a slightly lower average return during CPI Rising periods. This singular divergence may reflect higher sensitivity to sector-specific news.

## Investigation 7: Co-movement across all four assets

**Objective:** Count how often all four assets move in the same direction on the same day (all up or all down), to measure how correlated the portfolio really is versus how often the assets diverge.

```{r}
co_movement <- unified_clean |>
  mutate(
    all_positive = amd_return > 0 & eth_return > 0 & phi_return > 0 & spy_return > 0,
    all_negative = amd_return < 0 & eth_return < 0 & phi_return < 0 & spy_return < 0
  ) |>
  summarise(
    all_up_days = sum(all_positive, na.rm = TRUE),
    all_down_days = sum(all_negative, na.rm = TRUE),
    total_days = n(),
    pct_co_moving = (sum(all_positive, na.rm = TRUE) + sum(all_negative, na.rm = TRUE)) / n() * 100
  )

co_movement
```

Using the mutate() function, added columns to the unified database determine days where all assets are positive and negative, then summarised to only include those values and the percentage of movement. 

Out of 1,449 total trading days, all four assets moved in the same direction on only 385 days (212 all-up + 173 all-down), or roughly 26.6% of the time. This means that on the vast majority of trading days (~73%), the four assets did not move in unison, at least one asset diverged from the others.This implies high diversification of assets within our database.

## Investigation 8: Highest-volume trading days for AMD and ETH

**Objective:** Identify the 5 highest-volume trading days for AMD and ETH individually, since volume spikes often coincide with earnings releases, major news, or sharp price moves.

```{r}
amd_top_volume <- amd_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

eth_top_volume <- eth_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

amd_top_volume
eth_top_volume
```
Often times, volume spikes are associated with panic selling and liquidation of stocks, but all five of AMD's highest-volume days are strongly positive, with returns ranging from +5% to a striking +23.7% on 2025-10-06. This is likely consistent with AMD's earnings-release dates or announcements/news of AI/data-center chips, thus coinciding with the boom of AI and large language models. On the other hand, ETH's highest volume days present the opposite, four out of the five returns are negative, including a severe -27.2% crash on 2021-05-19 (this aligns with the May 2021 crypto market crash) and a more recent -12.0% drop on 2025-10-10. This supports the pattern where extreme trading volume in crypto markets tends to accompany sharp sell-offs and liquidation events.

## Investigation 9: Quarterly average price levels

**Objective:** Compute the average closing price per asset by quarter to see the longer-term price trend without daily noise.

```{r}
quarterly_prices <- unified_clean |>
  mutate(year_quarter = paste0(lubridate::year(date), "-Q", lubridate::quarter(date))) |>
  group_by(year_quarter) |>
  summarise(
    avg_spy_price = mean(spy_price, na.rm = TRUE),
    avg_amd_price = mean(amd_price, na.rm = TRUE),
    avg_phi_price = mean(phi_price, na.rm = TRUE),
    avg_eth_price = mean(eth_price, na.rm = TRUE)
  )

quarterly_prices
```

To assess an asset's long-term trend, closing prices are used as a representational basis for stock behavior, showing the pattern of risk and steadiness of the assets. Using the unified database, it is mutated to show the average closing stock prices of each asset per quartile of a year, and grouped by the yearly quartiles to determine a quarterly average.

Across 2020-Q1 through 2024-Q4, all four assets show a clear long-term upward trend, though at different magnitudes and different volatility patterns. SPY grows steadily and relatively smoothly, from ~\$277 (2020-Q1) to ~\$577 (2024-Q4), roughly a  2.1x growth. While AMD shows much sharper swings, climbing from ~\$48 to a peak of ~\$175 (2024-Q2), but with visible mid-cycle lows, such as falling from \$133 in 2021-Q4 to \$65 by 2022-Q4. PHI has the flattest growth, moving from ~\$13 to a modest ~\$22 by 2024-Q. Lastly, ETH proves to be the most volatile, from ~\$184 (2020-Q1) to a massive peak of ~\$4,096 (2021-Q4), then crashing to ~\$1,306 by 2022-Q4, before climbing again to ~\$3,055 by 2024-Q4, roughly a 22x increase from the start of the sample.

## Investigation 10: Risk-adjusted return comparison

**Objective:** Compute a simple Sharpe ratio for each asset (average daily return minus a daily risk-free rate, divided by return standard deviation) using the 10-Year Treasury yield as the risk-free proxy, to compare returns on a risk-adjusted basis rather than raw returns alone.

```{r}
risk_free_daily <- mean(unified_clean$treasury_10y, na.rm = TRUE) / 100 / 252

risk_adjusted <- unified_clean |>
  select(amd_return, eth_return, phi_return, spy_return) |>
  summarise(
    amd_sharpe = (mean(amd_return, na.rm = TRUE) - risk_free_daily) / sd(amd_return, na.rm = TRUE),
    eth_sharpe = (mean(eth_return, na.rm = TRUE) - risk_free_daily) / sd(eth_return, na.rm = TRUE),
    phi_sharpe = (mean(phi_return, na.rm = TRUE) - risk_free_daily) / sd(phi_return, na.rm = TRUE),
    spy_sharpe = (mean(spy_return, na.rm = TRUE) - risk_free_daily) / sd(spy_return, na.rm = TRUE)
  )

risk_adjusted
```

The Sharpe ratio measures an investment's return for each unit of risk taken, it helps in seeing if a higher return is due to smart investing or froim taking on too much risk. Even though ETH had the largest raw price appreciation over the period (see Investigation 9), SPY delivers a slightly better risk-adjusted return than ETH due to ETH's much higher volatility. AMD sits close behind both. The PHI asset, despite its steady price path in Investigation 9, actually has the weakest risk-adjusted return, its low volatility not proportional to any meaningful returns.

# Descriptive Statistics

From the previous section, a unified database named `unified_clean` was created — this is our final analytical dataset (post-join, post-fill, post-drop) that all statistics below are computed from.
```{r}
glimpse(unified_clean)
dim(unified_clean)
```

### 1. Selecting return variables
Since our end goal is an investment recommendation, we focus this analysis on the daily return series for each tradeable asset (return = risk/reward), while VIX and macro variables are analyzed separately as context in the Exploratory Financial Analysis section.
```{r}
returns_tbl <- unified_clean  |> 

  select(date,
         amd_return,
         eth_return,
         phi_return,
         spy_return)

head(returns_tbl)
dim(returns_tbl)

```

### 2. Computing median and mean
The general median and mean of returns don't give a good picture of how a "typical" return usually is due to outliers. It is just a simple average of the dataset. 
```{r}
#Mean Return
cat("AMD Mean: ")
mean(returns_tbl$amd_return,
     na.rm = TRUE)

#Median Return
cat("AMD Median: ")
median(returns_tbl$amd_return,
       na.rm = TRUE)

#Trimmed Mean
cat("AMD Trimmed Mean: ")
mean(returns_tbl$amd_return,
     trim = 0.05,
     na.rm = TRUE)

cat("ETH Mean: ")
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("ETH Median: ")
median(returns_tbl$eth_return,
       na.rm = TRUE)

cat("ETH Trimmed Mean: ")
mean(returns_tbl$eth_return,
     trim = 0.05,
     na.rm = TRUE)

cat("PHI Mean: ")
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("PHI Median: ")
median(returns_tbl$phi_return,
       na.rm = TRUE)

cat("PHI Trimmed Mean: ")
mean(returns_tbl$phi_return,
     trim = 0.05,
     na.rm = TRUE)

cat("SPY Mean: ")
mean(returns_tbl$spy_return,
     na.rm = TRUE)

cat("SPY Median: ")
median(returns_tbl$spy_return,
       na.rm = TRUE)

cat("SPY Trimmed Mean: ")
mean(returns_tbl$spy_return,
     trim = 0.05,
     na.rm = TRUE)

```

### 3. Computing Dispersion (Risk)
Calculating the variance, std dev, range, interquartile range, and coefficient of variation of returns shows a small view of each asset's general risk to return ratio.
```{r}
#Variance
cat("AMD Variance: ")
var(returns_tbl$amd_return,
    na.rm = TRUE)

#Standard Deviation
cat("AMD Standard Deviation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE)

#Range
cat("AMD Range: ")
range(returns_tbl$amd_return,
      na.rm = TRUE)

#Interquartile Range
cat("AMD IQR: ")
IQR(returns_tbl$amd_return,
    na.rm = TRUE)

#Coefficient of Variation
cat("AMD Coeff. of Variation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE) /
mean(returns_tbl$amd_return,
     na.rm = TRUE)

cat("ETH Variance: ")
var(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Standard Deviation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE)

cat("ETH Range: ")
range(returns_tbl$eth_return,
      na.rm = TRUE)

cat("ETH IQR: ")
IQR(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Coeff. of Variation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE) /
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("PHI Variance: ")
var(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Standard Deviation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE)

cat("PHI Range: ")
range(returns_tbl$phi_return,
      na.rm = TRUE)

cat("PHI IQR: ")
IQR(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Coeff. of Variation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE) /
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("SPY Variance: ")
var(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Standard Deviation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE)

cat("SPY Range: ")
range(returns_tbl$spy_return,
      na.rm = TRUE)

cat("SPY IQR: ")
IQR(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Coeff. of Variation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE) /
mean(returns_tbl$spy_return,
     na.rm = TRUE)

```

### 4. Viewing data shape
The skewness and kurtosis show the general shape of asset returns, where it is how often and how severely an asset's returns deviate from an averaged bell curve. Skewness measures the symmetry of returns, while kurtosis measures the frequency of extreme events.
```{r}
library(moments)
#Skewness
cat("AMD Skewness: ")
skewness(returns_tbl$amd_return,
         na.rm = TRUE)
#Kurtosis
cat("AMD Kurtosis: ")
kurtosis(returns_tbl$amd_return,
         na.rm = TRUE)

cat("ETH Skewness: ")
skewness(returns_tbl$eth_return,
         na.rm = TRUE)

cat("ETH Kurtosis: ")
kurtosis(returns_tbl$eth_return,
         na.rm = TRUE)

cat("PHI Skewness: ")
skewness(returns_tbl$phi_return,
         na.rm = TRUE)

cat("PHI Kurtosis: ")
kurtosis(returns_tbl$phi_return,
         na.rm = TRUE)

cat("SPY Skewness: ")
skewness(returns_tbl$spy_return,
         na.rm = TRUE)

cat("SPY Kurtosis: ")
kurtosis(returns_tbl$spy_return,
         na.rm = TRUE)
```

### 5. Computing relative data position
We calculate the minimum and maximum of each asset's return shows the biggest drop and the largest jump in value over a set time. While quartiles and percentiles of asset returns divide an asset's past returns into parts to show how often it performs well or poorly. These help in seeing an investment's risk and reward at a glance.
```{r}
#Minimum
cat("AMD Minimum: ")
min(returns_tbl$amd_return,
    na.rm = TRUE)

#Maximum
cat("AMD Maximum: ")
max(returns_tbl$amd_return,
    na.rm = TRUE)

#Quartiles
cat("AMD Quartiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

#Percentiles
cat("AMD Percentiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("ETH Minimum: ")
min(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Maximum: ")
max(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Quartiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("ETH Percentiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("PHI Minimum: ")
min(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Maximum: ")
max(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Quartiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("PHI Percentiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("SPY Minimum: ")
min(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Maximum: ")
max(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Quartiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("SPY Percentiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)
```

### 6. Dealing with Missing Values
Usually datasets have null rows that can affect the general distribuition of statistical analysis, thus null rows are to be dropped or imputed depending on the information held.
```{r}
colSums(is.na(returns_tbl))
```
Considering no null rows exist in our dataset we proceed without having to drop or impute.

### 7. Correlation Matrix
A correlation matrix of asset returns measures how investments move together and is a standardized form of the covariance that normalizes into a score between \(-1\) and \(1\).
```{r}
correlation_matrix <- returns_tbl  |> 
  select(-date)  |> 
  cor(use = "complete.obs")
correlation_matrix
```

### 8. Covariance Matrix 
A covariance matrix of asset returns also measures how investments move together,it shows the direction of the relationship between pairs of assets 
```{r}
covariance_matrix <- returns_tbl |> 
  select(-date)  |> 
  cov(use = "complete.obs")
covariance_matrix
```

### 9. Final Descriptive Statistics Table
```{r}
library(psych)

desc_stats <- psych::describe(
  returns_tbl  |> 
    select(-date)
)
round(desc_stats,4)
```

```{r}
descriptive_table <- tibble(

  Statistic = c(
    "Mean",
    "Median",
    "Trimmed Mean",
    "Standard Deviation",
    "Variance",
    "Minimum",
    "Q1",
    "Q3",
    "Maximum",
    "Range",
    "Interquartile Range",
    "Coefficient of Variation",
    "Skewness",
    "Kurtosis",
    "Observations"
  ),

  AMD = c(
    mean(returns_tbl$amd_return, na.rm = TRUE),
    median(returns_tbl$amd_return, na.rm = TRUE),
    mean(returns_tbl$amd_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE),
    var(returns_tbl$amd_return, na.rm = TRUE),
    min(returns_tbl$amd_return, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.75, na.rm = TRUE),
    max(returns_tbl$amd_return, na.rm = TRUE),
    diff(range(returns_tbl$amd_return, na.rm = TRUE)),
    IQR(returns_tbl$amd_return, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE) /
      mean(returns_tbl$amd_return, na.rm = TRUE),
    skewness(returns_tbl$amd_return, na.rm = TRUE),
    kurtosis(returns_tbl$amd_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$amd_return))
  ),
  
  ETH = c(
    mean(returns_tbl$eth_return, na.rm = TRUE),
    median(returns_tbl$eth_return, na.rm = TRUE),
    mean(returns_tbl$eth_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE),
    var(returns_tbl$eth_return, na.rm = TRUE),
    min(returns_tbl$eth_return, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.75, na.rm = TRUE),
    max(returns_tbl$eth_return, na.rm = TRUE),
    diff(range(returns_tbl$eth_return, na.rm = TRUE)),
    IQR(returns_tbl$eth_return, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE) /
      mean(returns_tbl$eth_return, na.rm = TRUE),
    skewness(returns_tbl$eth_return, na.rm = TRUE),
    kurtosis(returns_tbl$eth_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$eth_return))
  ),
  
  PHI = c(
    mean(returns_tbl$phi_return, na.rm = TRUE),
    median(returns_tbl$phi_return, na.rm = TRUE),
    mean(returns_tbl$phi_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE),
    var(returns_tbl$phi_return, na.rm = TRUE),
    min(returns_tbl$phi_return, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.75, na.rm = TRUE),
    max(returns_tbl$phi_return, na.rm = TRUE),
    diff(range(returns_tbl$phi_return, na.rm = TRUE)),
    IQR(returns_tbl$phi_return, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE) /
      mean(returns_tbl$phi_return, na.rm = TRUE),
    skewness(returns_tbl$phi_return, na.rm = TRUE),
    kurtosis(returns_tbl$phi_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$phi_return))
  ),
  
  SPY = c(
    mean(returns_tbl$spy_return, na.rm = TRUE),
    median(returns_tbl$spy_return, na.rm = TRUE),
    mean(returns_tbl$spy_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE),
    var(returns_tbl$spy_return, na.rm = TRUE),
    min(returns_tbl$spy_return, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.75, na.rm = TRUE),
    max(returns_tbl$spy_return, na.rm = TRUE),
    diff(range(returns_tbl$spy_return, na.rm = TRUE)),
    IQR(returns_tbl$spy_return, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE) /
      mean(returns_tbl$spy_return, na.rm = TRUE),
    skewness(returns_tbl$spy_return, na.rm = TRUE),
    kurtosis(returns_tbl$spy_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$spy_return))
  )
)

descriptive_table


```
# Visualization

All figures below are built from `unified_clean`, our final analytical dataset, so they stay consistent with every statistic and SQL investigation computed earlier.

## Figure 1: Indexed Price Trend (Base = 100)

Raw prices aren't directly comparable since SPY, AMD, PHI, and ETH trade on completely different price scales. Rebasing each series to start at 100 lets us compare growth trajectories side by side.
```{r}
indexed_prices <- unified_clean |>
  select(date, spy_price, amd_price, phi_price, eth_price) |>
  mutate(across(-date, ~ .x / first(.x) * 100)) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "indexed_price")

ggplot(indexed_prices, aes(x = date, y = indexed_price, color = asset)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Figure 1: Indexed Price Trend (Base = 100, Jan 2020)",
    x = "Date",
    y = "Indexed Price"
  ) +
  theme_minimal()
```

**Interpretation:** This chart is shows which asset had the strongest overall run across the full sample period and which one was the most volatile along the way, since a shaky line with a high endpoint (ETH) tells a very different investment story than a steady line with the same endpoint.

## Figure 2: Distribution of Daily Returns by Asset

```{r}
returns_long_viz <- unified_clean |>
  select(date, amd_return, eth_return, phi_return, spy_return) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "return")

ggplot(returns_long_viz, aes(x = return, fill = asset)) +
  geom_histogram(bins = 60) +
  facet_wrap(~asset, scales = "free") +
  labs(
    title = "Figure 2: Distribution of Daily Returns by Asset",
    x = "Daily Return",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation:** This is a direct Anscombe's Quartet check for return data. Ttwo assets can have nearly identical mean and standard deviation on paper while looking completely different in terms of visualization, one might be a tight, roughly normal bell curve while another has fat tails or a lopsided shape. But, as seen across all assets, they more or less follow a binomial distribution.

## Figure 3: 30-Day Rolling Volatility by Asset

```{r}
rolling_vol <- unified_clean |>
  arrange(date) |>
  mutate(
    amd_vol_30d = zoo::rollapply(amd_return, width = 30, FUN = sd, fill = NA, align = "right"),
    eth_vol_30d = zoo::rollapply(eth_return, width = 30, FUN = sd, fill = NA, align = "right"),
    phi_vol_30d = zoo::rollapply(phi_return, width = 30, FUN = sd, fill = NA, align = "right"),
    spy_vol_30d = zoo::rollapply(spy_return, width = 30, FUN = sd, fill = NA, align = "right")
  ) |>
  select(date, amd_vol_30d, eth_vol_30d, phi_vol_30d, spy_vol_30d) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "rolling_vol")

ggplot(rolling_vol, aes(x = date, y = rolling_vol, color = asset)) +
  geom_line(linewidth = 0.6, na.rm = TRUE) +
  labs(
    title = "Figure 3: 30-Day Rolling Volatility by Asset",
    x = "Date",
    y = "30-Day Rolling Std. Dev. of Return"
  ) +
  theme_minimal()
```

**Interpretation:** A static standard deviation number treats risk as constant across the whole 2020-2025 window. Certain spikes line up across all four lines at once, this can point to a market-wide event (like a rate decision or macro shock). This also relates to the High VIX vs Low VIX comparison from the SQL Investigation section by showing exactly when volatility regimes shifted.

## Figure 4: SPY Daily Return vs VIX Level

```{r}
ggplot(unified_clean, aes(x = vix, y = spy_return)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(
    title = "Figure 4: SPY Daily Return vs VIX Level",
    x = "VIX Level",
    y = "SPY Daily Return"
  ) +
  theme_minimal()
```

**Interpretation:** Since the fitted line slopes downward, higher fear/uncertainty (elevated VIX) coincides with weaker SPY returns, confirming VIX's role as a risk signal or a "fear gauge". The spread of points around the line also matters: a wide scatter at high VIX levels shows returns get less predictable exactly when the market is most stressed.

## Figure 5: Correlation Heatmap of Daily Returns

```{r}
corr_df <- as.data.frame(correlation_matrix) |>
  tibble::rownames_to_column("asset1") |>
  tidyr::pivot_longer(-asset1, names_to = "asset2", values_to = "correlation")

ggplot(corr_df, aes(x = asset1, y = asset2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", midpoint = 0) +
  labs(
    title = "Figure 5: Correlation Heatmap of Daily Returns",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()
```

**Interpretation:** This is tests the diversification of our assets for the final allocation. Values near 0 between two assets mean they largely move independently, while values close to 1 mean the assets are effectively matching each other's risk, so holding both doesn't reduce overall exposure the way holding two lowly correlated assets would.

## Figure 6: Cumulative Growth of $1 Invested

```{r}
cumulative_growth <- unified_clean |>
  arrange(date) |>
  mutate(
    amd_cum = cumprod(1 + coalesce(amd_return, 0)),
    eth_cum = cumprod(1 + coalesce(eth_return, 0)),
    phi_cum = cumprod(1 + coalesce(phi_return, 0)),
    spy_cum = cumprod(1 + coalesce(spy_return, 0))
  ) |>
  select(date, amd_cum, eth_cum, phi_cum, spy_cum) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "growth_of_1")

ggplot(cumulative_growth, aes(x = date, y = growth_of_1, color = asset)) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  labs(
    title = "Figure 6: Cumulative Growth of $1 Invested (2020-2025)",
    x = "Date",
    y = "Value of $1 Invested"
  ) +
  theme_minimal()
```

**Interpretation:** This is a compounding view rather than a daily-return view, small consistent gains stack up very differently than occasional large spikes. An asset that looked flashy in Figure 2's histogram might still end up behind a steadier asset here once compounding and drawdowns are accounted for. This chart is a good direct visual to pair with the risk-adjusted return numbers from SQL Investigation 10.

## Figure 7: Daily Return Distribution by Asset (Boxplot)

```{r}
ggplot(returns_long_viz, aes(x = asset, y = return, fill = asset)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Figure 7: Daily Return Distribution by Asset",
    x = "Asset",
    y = "Daily Return"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation:** Boxplots make the risk comparison from the descriptive statistics table visual, box height reflects the interquartile range (typical-day risk) while the whiskers and outlier points show how extreme the worst days actually got. An asset with a tight box but long outlier whiskers has low everyday risk but real tail risk.

## Figure 8: SPY Price vs CPI and Fed Funds Rate Over Time

```{r}
macro_long <- unified_clean |>
  select(date, spy_price, cpi, fedfunds) |>
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value")

ggplot(macro_long, aes(x = date, y = value)) +
  geom_line(color = "darkslategray", na.rm = TRUE) +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(
    title = "Figure 8: SPY Price vs CPI and Fed Funds Rate Over Time",
    x = "Date",
    y = NULL
  ) +
  theme_minimal()
```

**Interpretation:** Stacking these on a shared date axis but separate scales lets us visually line up SPY's price trend against the broader macroeconomic environment Look for whether SPY's biggest run-ups or drawdowns line up with periods of aggressive Fed Funds hikes or accelerating CPI.

# Exploratory Financial Analysis

This section does not recompute new statistics from scratch. It pulls together what was already built in the SQL Investigation and Visualization sections and interprets them together, since the spec asks us to compare across five specific angles: returns, volatility, inflationary periods, high-VIX periods, and interest-rate environments.

## 1. Comparing Returns and Volatility Across Assets

```{r}
return_vol_summary <- returns_long_viz |>
  group_by(asset) |>
  summarise(
    avg_daily_return = mean(return, na.rm = TRUE),
    annualized_return = mean(return, na.rm = TRUE) * 252,
    daily_volatility = sd(return, na.rm = TRUE),
    annualized_volatility = sd(return, na.rm = TRUE) * sqrt(252)
  ) |>
  arrange(desc(annualized_return))

return_vol_summary
```

Annualizing the daily figures for 252 trading days puts all four assets on a comparable yearly basis instead of the tiny daily numbers that are hard to reason about directly. Pairing this table with Figure 1 (indexed price trend) and Figure 6 (cumulative growth of $1), an asset can have a strong annualized return on paper while Figure 6 shows the path there was rough with deep drawdowns along the way. Figure 7's boxplot is the fastest way to see this same return-vs-volatility tradeoff visually, a tall box means more day-to-day risk for whatever return the asset ultimately delivered.



## 2. Inflationary Periods

```{r}
cpi_trend
```

**Interpretation:** This reuses the CPI-rising vs CPI-flat/falling split from SQL Investigation 6. Since CPI only updates monthly and we forward-filled it across each trading day, "CPI Rising" here means the whole month following a higher print, not a single release-day reaction. Cross-reference this with Figure 8, where the CPI panel shows the actual inflation path over 2020-2025, to see whether the rising-CPI periods in the table line up with the sharpest sections of that line. If equity and crypto returns are consistently weaker during CPI Rising months, that is direct evidence inflation surprises pressure risk assets in this dataset, not just a theoretical assumption.

## 3. High-VIX Periods

```{r}
vix_regime_stats
```

**Interpretation:** This reuses the High VIX vs Low VIX split from SQL Investigation 4, built around the median VIX level in our sample. Figure 3 (30-day rolling volatility) and Figure 4 (SPY return vs VIX scatter) are the visual backbone for this comparison, Figure 3 shows exactly when the market moved into a High VIX regime, and Figure 4 shows whether higher VIX levels actually coincided with weaker or more unpredictable SPY returns on a day-by-day basis rather than just in aggregate. If both the standard deviation and the average return move in the expected direction (higher volatility, lower average return) during High VIX days, that is a strong basis for treating VIX as an early warning signal in the investment recommendation, not just a descriptive curiosity.

## 4. Interest-Rate Environments

```{r}
fedfunds_trend
```

**Interpretation:** This reuses the Rate Hike vs Rate Cut vs Held Steady split from SQL Investigation 5. Crypto is generally the most sensitive of the four assets to rate changes, since it carries no yield of its own to compete with a higher risk-free rate, so ETH's average return during Rate Hike days is worth checking closely against AMD, PHI, and SPY. Figure 8's Fed Funds panel shows the actual rate path across the sample, use it to see whether the current environment (as of the end of the sample period) sits in a hiking, cutting, or holding phase, since that context should carry directly into how much crypto exposure the investment recommendation allocates going forward.

## 5. Bringing the Comparisons Together

**Interpretation:** Across all four lenses, average and annualized returns (Section 1), inflationary periods (Section 2), high-VIX periods (Section 3), and interest-rate environments (Section 4), the goal is to identify which asset holds up best under stress (High VIX, CPI Rising, Rate Hike) rather than only which asset has the best raw average return in calm conditions. An asset that leads on annualized return but falls apart hardest under all three stress conditions is a very different allocation decision than one that trails slightly on raw return but stays comparatively stable. This is the evidence base the Investment Recommendation section should draw from directly, rather than restating the descriptive statistics table in isolation.

# Investment Recommendation
"Recommend allocation of USD 1,000,000 among the selected U.S. asset, Philippine asset, cryptocurrency and cash." -- from the specs
```{r}
```

# References
- booyah