---
title: "FINSTAD Financial Database"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
```

## Comparing the different datasets
note: we probably need to explicitly write down and explain each set smt like that

We first get symbols from the following sources:

- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis' FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

First we can look at the macroeconomic data

The data is taken from the Federal Reserve Economic Database (FRED) from the Federal Reserve Bank of St. Louis, where we load in the _Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity_ (**DGS10**) and _Consumer Price Index for All Urban Consumers_ (**CPIAUCSL**).

**DGS10** is used as a proxy for the "risk-free rate" in finance, it helps investors gauge inflation expectations, economic growth, and monetary policy shifts. While **CPIAUCSL** is seasonally adjusted to gauge monthly inflation metrics on consumer goods.

These pieces of macroeconomic data would help in
```{r}
head(DGS10)
head(CPIAUCSL)
```

We can then take a look at the stock data
```{r}
ETH <- `ETH-USD`

head(AMD)
head(PHI)
head(ETH)
```

Now we take a look at VIX and SPY data
```{r}
head(SPY)
head(VIX)
```

# Cleaning

## 1. Converting XTS to tibbles
First we convert all XTS objects to tibbles.(expound)
```{r}
dgs10_tbl <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl_tbl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy_tbl <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
  adj_close = as.numeric(Ad(SPY)),
)
amd_tbl <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
  adj_close = as.numeric(Ad(AMD)),
)
phi_tbl <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
  adj_close = as.numeric(Ad(PHI)),
)
vix_tbl <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
  adj_close = as.numeric(Ad(VIX)),
)
eth_tbl <- tibble(
  date = index(ETH),
  open = as.numeric(ETH$`ETH-USD.Open`),
  high = as.numeric(ETH$`ETH-USD.High`),
  low = as.numeric(ETH$`ETH-USD.Low`),
  close = as.numeric(ETH$`ETH-USD.Close`),
  volume = as.numeric(ETH$`ETH-USD.Volume`),
  adjusted = as.numeric(ETH$`ETH-USD.Adjusted`),
  adj_close = as.numeric(Ad(ETH)),
)
```

## 2. Asset Identifier
We add an asset identifier per asset to smt smt
```{r}
amd_tbl <- amd_tbl |> 
  mutate(asset = "AMD")

eth_tbl <- eth_tbl |> 
  mutate(asset = "ETH")

phi_tbl <- phi_tbl |> 
  mutate(asset = "PHI")

spy_tbl <- spy_tbl |> 
  mutate(asset = "SPY")

vix_tbl <- vix_tbl |> 
  mutate(asset = "VIX")
```

## 3. Computing daily returns
idk if we should compute basic returns here or log returns
```{r}
amd_tbl <- amd_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

eth_tbl <- eth_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

phi_tbl <- phi_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

spy_tbl <- spy_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

vix_tbl <- vix_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )
```

## 4. Removing missing rows
idk if we should impute or just remove null rows 
```{r}
amd_tbl <- amd_tbl  |> 
  filter(!is.na(return))

eth_tbl <- eth_tbl  |> 
  filter(!is.na(return))

phi_tbl <- phi_tbl |> 
  filter(!is.na(return))

spy_tbl <- spy_tbl  |> 
  filter(!is.na(return))

vix_tbl <- vix_tbl  |> 
  filter(!is.na(return))
```

## 5. Examine the data
```{r}
glimpse(amd_tbl)
glimpse(eth_tbl)
glimpse(phi_tbl)
glimpse(spy_tbl)
glimpse(vix_tbl)
```

Summary Tables
```{r}
summary(amd_tbl)
summary(eth_tbl)
summary(phi_tbl)
summary(spy_tbl)
summary(vix_tbl)
```

## 6. Preparing Macroeconomic variables
```{r}
treasury_tbl <- tibble(
  date = index(DGS10),
  treasury_10y = as.numeric(DGS10$DGS10)
)
```

```{r}
cpi_tbl <- tibble(
  date = index(CPIAUCSL),
  cpi = as.numeric(CPIAUCSL$CPIAUCSL)
)
```

```{r}
head(treasury_tbl)
head(cpi_tbl)
```

## 7. Create an Observation Summary Table and saving it
We first count the observations
```{r}
nrow(amd_tbl)
nrow(eth_tbl)
nrow(phi_tbl)
nrow(spy_tbl)
nrow(vix_tbl)
nrow(treasury_tbl)
nrow(cpi_tbl)
```

Creating the observation summary table
```{r}
obs_summary <- tibble(
  Dataset = c("AMD",
              "Ethereum",
              "PLDT",
              "SPY",
              "VIX",
              "Treasury",
              "CPI"),
  Observations = c(
    nrow(amd_tbl),
    nrow(eth_tbl),
    nrow(phi_tbl),
    nrow(spy_tbl),
    nrow(vix_tbl),
    nrow(treasury_tbl),
    nrow(cpi_tbl)
  )
)

obs_summary
```

Saving the clean tables
```{r}
amd_clean <- amd_tbl
eth_clean <- eth_tbl
phi_clean <- phi_tbl
spy_clean <- spy_tbl
vix_clean <- vix_tbl
treasury_clean <- treasury_tbl
cpi_clean <- cpi_tbl
```

# SQL Investigation
"Investigate at least 10 meaningful financial questions using select(), filter(), mutate(), arrange(), group_by() and summarise(). Each query must include Objective, Code, Output and Interpretation." -- from the specs

After cleaning all the data, we collate the rows into 1 unified dataset ? idk if thats correct wording

## 1. Renaming variables
```{r}
```

## 2. from this number till the "Compare joins" number we need to enumerate why we used a certain join on what data and how many

## <num0>. Compare joins
```{r}
```

## <num1>. Building the Unified Database
```{r}
```

## <num2>. Observation Tracking
```{r}
```

## <num3>. Dropping Null Rows
```{r}
```

# Descriptive Statistics
"Compute mean, median, variance, standard deviation, minimum, maximum, range, skewness and kurtosis." -- from the specs

From the previous section, a unified database named `<db name>` was created, we can first examine the variables
```{r}
```

## 1. Selecting return variables
ok im assuming since our end goal is an investment strategy, we are to analyze the returns + voilatility?? idk
```{r}
```

## 2. Computing median and mean
```{r}
```

## 3. Computing Dispersion (Risk)
variance, std dev, range, interquartile range, coefficient of variation
```{r}
```

## 4. Viewing data shape
skewness and kurtosis
```{r}
```

## 5. Computing data position ?
min, max, quartiles, percentiles
```{r}
```

## 6. Dealing with Missing Values
```{r}
```

## 7. Correlation Matrix
```{r}
```

## 8. Covariance Matrix 
```{r}
```

## 9. Final Descriptive Statistics Table
```{r}
library(psych)
```

# Visualization
"Produce at least eight professional figures with interpretations." -- from the specs

```{r}
```

(for these sections below im not sure if this should be in the paper or in the RMD file itself, or we place it here then elaborate further in the paper)
# Exploratory Financial Analysis
"Compare returns, volatility, inflationary periods, high-VIX periods and interest-rate environments." -- from the specs
```{r}
```

# Investment Recommendation
"Recommend allocation of USD 1,000,000 among the selected U.S. asset, Philippine asset, cryptocurrency and cash." -- from the specs
```{r}
```

# References
- booyah