---
title: "FINSTAD Financial Database"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
```

### Comparing the different datasets
note: we probably need to explicitly write down and explain each set smt like that

We first get symbols from the following sources:
- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis' FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

Then we can look at the macroeconomic data,

The data is taken from the Federal Reserve Economic Database (FRED) from the Federal Reserve Bank of St. Louis, where we load in the _Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity_ (**DGS10**) and _Consumer Price Index for All Urban Consumers_ (**CPIAUCSL**).

**DGS10** is used as a proxy for the "risk-free rate" in finance, it helps investors gauge inflation expectations, economic growth, and monetary policy shifts. While **CPIAUCSL** is seasonally adjusted to gauge monthly inflation metrics on consumer goods.

These pieces of macroeconomic data would help in understanding broader market cycles and how systemic risks affect stock valuations.
```{r}
head(DGS10)
head(CPIAUCSL)
```

We can then take a look at our chosen financial market data,

U.S. Asset - Advanced Micro Devices, Inc. (AMD)
Philippine Asset - PLDT, Inc.(PHI)
Cryptocurrency - Etherium (ETH or ETH-USD)
Required - State Street SPDR S&P 500 ETF Trust (SPY)

*note: do we explain why we chose these companies lol??
```{r}
ETH <- `ETH-USD`

head(AMD)
head(PHI)
head(ETH)
head(SPY)
```

Lastly, we take a look at the _Cboe Volatility Index_ (**VIX**), this is not an individual stock but a real time market index created by the Chicago Board Options Exchange (Cboe). This measures the market's expectation of 30-day volatility in the S&P 500 index.

This adds aid on top of macroeconomic data to further gauge systemic risks, VIX is known as the "Fear Gauge" which reflects levels of anxiety, complacency, or uncertainty in the market, when this "Fear Gauge" level rises it implies larger expected swings in stock prices.
```{r}
head(VIX)
```

# Cleaning

### 1. Converting XTS to tibbles
First we convert all XTS objects to tibbles for easier data manipulation.
```{r}
dgs10_tbl <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl_tbl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy_tbl <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
  adj_close = as.numeric(Ad(SPY)),
)
amd_tbl <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
  adj_close = as.numeric(Ad(AMD)),
)
phi_tbl <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
  adj_close = as.numeric(Ad(PHI)),
)
vix_tbl <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
  adj_close = as.numeric(Ad(VIX)),
)
eth_tbl <- tibble(
  date = index(ETH),
  open = as.numeric(ETH$`ETH-USD.Open`),
  high = as.numeric(ETH$`ETH-USD.High`),
  low = as.numeric(ETH$`ETH-USD.Low`),
  close = as.numeric(ETH$`ETH-USD.Close`),
  volume = as.numeric(ETH$`ETH-USD.Volume`),
  adjusted = as.numeric(ETH$`ETH-USD.Adjusted`),
  adj_close = as.numeric(Ad(ETH)),
)
```

### 2. Asset Identifier
We add an asset identifier per asset to see the data better in table structures.
```{r}
amd_tbl <- amd_tbl |> 
  mutate(asset = "AMD")

eth_tbl <- eth_tbl |> 
  mutate(asset = "ETH")

phi_tbl <- phi_tbl |> 
  mutate(asset = "PHI")

spy_tbl <- spy_tbl |> 
  mutate(asset = "SPY")

vix_tbl <- vix_tbl |> 
  mutate(asset = "VIX")
```

### 3. Computing daily returns
idk if we should compute basic returns here or log returns

Daily returns are computed using the basic return formula
```{r}
amd_tbl <- amd_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

eth_tbl <- eth_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

phi_tbl <- phi_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

spy_tbl <- spy_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

vix_tbl <- vix_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )
```

### 4. Removing missing rows
idk if we should impute or just remove null rows 
```{r}
amd_tbl <- amd_tbl  |> 
  filter(!is.na(return))

eth_tbl <- eth_tbl  |> 
  filter(!is.na(return))

phi_tbl <- phi_tbl |> 
  filter(!is.na(return))

spy_tbl <- spy_tbl  |> 
  filter(!is.na(return))

vix_tbl <- vix_tbl  |> 
  filter(!is.na(return))
```

### 5. Examine the data
```{r}
glimpse(amd_tbl)
glimpse(eth_tbl)
glimpse(phi_tbl)
glimpse(spy_tbl)
glimpse(vix_tbl)
```

Summary Tables
```{r}
summary(amd_tbl)
summary(eth_tbl)
summary(phi_tbl)
summary(spy_tbl)
summary(vix_tbl)
```

### 6. Preparing Macroeconomic variables
```{r}
treasury_tbl <- tibble(
  date = index(DGS10),
  treasury_10y = as.numeric(DGS10$DGS10)
)
```

```{r}
cpi_tbl <- tibble(
  date = index(CPIAUCSL),
  cpi = as.numeric(CPIAUCSL$CPIAUCSL)
)
```

```{r}
head(treasury_tbl)
head(cpi_tbl)
```

### 7. Create an Observation Summary Table and saving it
We first count the observations
```{r}
nrow(amd_tbl)
nrow(eth_tbl)
nrow(phi_tbl)
nrow(spy_tbl)
nrow(vix_tbl)
nrow(treasury_tbl)
nrow(cpi_tbl)
```

Creating the observation summary table
```{r}
obs_summary <- tibble(
  Dataset = c("AMD",
              "Ethereum",
              "PLDT",
              "SPY",
              "VIX",
              "Treasury",
              "CPI"),
  Observations = c(
    nrow(amd_tbl),
    nrow(eth_tbl),
    nrow(phi_tbl),
    nrow(spy_tbl),
    nrow(vix_tbl),
    nrow(treasury_tbl),
    nrow(cpi_tbl)
  )
)

obs_summary
```

Saving the clean tables
```{r}
amd_clean <- amd_tbl
eth_clean <- eth_tbl
phi_clean <- phi_tbl
spy_clean <- spy_tbl
vix_clean <- vix_tbl
treasury_clean <- treasury_tbl
cpi_clean <- cpi_tbl
```

# SQL Investigation
"Investigate at least 10 meaningful financial questions using select(), filter(), mutate(), arrange(), group_by() and summarise(). Each query must include Objective, Code, Output and Interpretation." -- from the specs

After cleaning all the data, we collate the rows into 1 unified dataset.

## 1. Renaming variables
```{r}
amd_sql <- amd_clean  |> 
  select(date,
         amd_price = adj_close,
         amd_return = return)

eth_sql <- eth_clean  |> 
  select(date,
         eth_price = adj_close,
         eth_return = return)

phi_sql <- phi_clean  |> 
  select(date,
         phi_price = adj_close,
         phi_return = return)

spy_sql <- spy_clean  |> 
  select(date,
         spy_price = adj_close,
         spy_return = return)
```

```{r}
head(amd_sql)
head(eth_sql)
head(phi_sql)
head(spy_sql)
```

### 2. from this number till the "Compare joins" number we need to enumerate why we used a certain join on what data and how many
```{r}

```

### <num0>. Compare joins
```{r}
```

### <num1>. Building the Unified Database
```{r}
unified_db <- spy_sql |> 
  left_join(amd_sql,
            by = "date") |> 
  left_join(phi_sql,
            by = "date") |>
  left_join(eth_sql,
            by = "date") |> 
  left_join(vix_clean |> 
              select(date, 
                     vix = adj_close,
                     vix_return = return),
            by = "date") |> 
  left_join(treasury_clean,
            by = "date") |> 
  left_join(cpi_clean,
            by = "date")
  #arrange(date) |> 
  #tidyr::fill(treasury_10y, 
  #            cpi, 
  #            .direction = "down")

glimpse(unified_db)
head(unified_db)
view(unified_db)
```

### <num2>. Observation Tracking
```{r}
```

Checking null rows
```{r}
colSums(
  is.na(
    unified_db
  )
)
```

### <num3>. Dropping Null Rows
```{r}
```

# Descriptive Statistics
"Compute mean, median, variance, standard deviation, minimum, maximum, range, skewness and kurtosis." -- from the specs

From the previous section, a unified database named `<db name>` was created, we can first examine the variables
```{r}

```

### 1. Selecting return variables
ok im assuming since our end goal is an investment strategy, we are to analyze the returns + voilatility?? idk
```{r}
returns_tbl <- unified_db  |> 

  select(date,
         amd_return,
         eth_return,
         phi_return,
         spy_return)

head(returns_tbl)
dim(returns_tbl)

```

### 2. Computing median and mean
```{r}
#Mean Return
cat("AMD Mean: ")
mean(returns_tbl$amd_return,
     na.rm = TRUE)

#Median Return
cat("AMD Median: ")
median(returns_tbl$amd_return,
       na.rm = TRUE)

#Trimmed Mean
cat("AMD Trimmed Mean: ")
mean(returns_tbl$amd_return,
     trim = 0.05,
     na.rm = TRUE)

cat("ETH Mean: ")
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("ETH Median: ")
median(returns_tbl$eth_return,
       na.rm = TRUE)

cat("ETH Trimmed Mean: ")
mean(returns_tbl$eth_return,
     trim = 0.05,
     na.rm = TRUE)

cat("PHI Mean: ")
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("PHI Median: ")
median(returns_tbl$phi_return,
       na.rm = TRUE)

cat("PHI Trimmed Mean: ")
mean(returns_tbl$phi_return,
     trim = 0.05,
     na.rm = TRUE)

cat("SPY Mean: ")
mean(returns_tbl$spy_return,
     na.rm = TRUE)

cat("SPY Median: ")
median(returns_tbl$spy_return,
       na.rm = TRUE)

cat("SPY Trimmed Mean: ")
mean(returns_tbl$spy_return,
     trim = 0.05,
     na.rm = TRUE)

```

### 3. Computing Dispersion (Risk)
variance, std dev, range, interquartile range, coefficient of variation
```{r}
#Variance
cat("AMD Variance: ")
var(returns_tbl$amd_return,
    na.rm = TRUE)

#Standard Deviation
cat("AMD Standard Deviation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE)

#Range
cat("AMD Range: ")
range(returns_tbl$amd_return,
      na.rm = TRUE)

#Interquartile Range
cat("AMD IQR: ")
IQR(returns_tbl$amd_return,
    na.rm = TRUE)

#Coefficient of Variation
cat("AMD Coeff. of Variation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE) /
mean(returns_tbl$amd_return,
     na.rm = TRUE)

cat("ETH Variance: ")
var(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Standard Deviation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE)

cat("ETH Range: ")
range(returns_tbl$eth_return,
      na.rm = TRUE)

cat("ETH IQR: ")
IQR(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Coeff. of Variation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE) /
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("PHI Variance: ")
var(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Standard Deviation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE)

cat("PHI Range: ")
range(returns_tbl$phi_return,
      na.rm = TRUE)

cat("PHI IQR: ")
IQR(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Coeff. of Variation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE) /
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("SPY Variance: ")
var(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Standard Deviation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE)

cat("SPY Range: ")
range(returns_tbl$spy_return,
      na.rm = TRUE)

cat("SPY IQR: ")
IQR(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Coeff. of Variation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE) /
mean(returns_tbl$spy_return,
     na.rm = TRUE)

```

### 4. Viewing data shape
skewness and kurtosis
```{r}
library(moments)
#Skewness
cat("AMD Skewness: ")
skewness(returns_tbl$amd_return,
         na.rm = TRUE)
#Kurtosis
cat("AMD Kurtosis: ")
kurtosis(returns_tbl$amd_return,
         na.rm = TRUE)

cat("ETH Skewness: ")
skewness(returns_tbl$eth_return,
         na.rm = TRUE)

cat("ETH Kurtosis: ")
kurtosis(returns_tbl$eth_return,
         na.rm = TRUE)

cat("PHI Skewness: ")
skewness(returns_tbl$phi_return,
         na.rm = TRUE)

cat("PHI Kurtosis: ")
kurtosis(returns_tbl$phi_return,
         na.rm = TRUE)

cat("SPY Skewness: ")
skewness(returns_tbl$spy_return,
         na.rm = TRUE)

cat("SPY Kurtosis: ")
kurtosis(returns_tbl$spy_return,
         na.rm = TRUE)
```

### 5. Computing data position ?
min, max, quartiles, percentiles
```{r}
#Minimum
cat("AMD Minimum: ")
min(returns_tbl$amd_return,
    na.rm = TRUE)

#Maximum
cat("AMD Maximum: ")
max(returns_tbl$amd_return,
    na.rm = TRUE)

#Quartiles
cat("AMD Quartiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

#Percentiles
cat("AMD Percentiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("ETH Minimum: ")
min(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Maximum: ")
max(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Quartiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("ETH Percentiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("PHI Minimum: ")
min(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Maximum: ")
max(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Quartiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("PHI Percentiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("SPY Minimum: ")
min(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Maximum: ")
max(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Quartiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("SPY Percentiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)
```

### 6. Dealing with Missing Values
```{r}
colSums(is.na(returns_tbl))
```

### 7. Correlation Matrix
```{r}
correlation_matrix <- returns_tbl  |> 
  select(-date)  |> 
  cor(use = "complete.obs")
correlation_matrix
```

### 8. Covariance Matrix 
```{r}
covariance_matrix <- returns_tbl |> 
  select(-date)  |> 
  cov(use = "complete.obs")
covariance_matrix
```

### 9. Final Descriptive Statistics Table
```{r}
library(psych)

desc_stats <- psych::describe(
  returns_tbl  |> 
    select(-date)
)
round(desc_stats,4)
```

```{r}
descriptive_table <- tibble(

  Statistic = c(
    "Mean",
    "Median",
    "Trimmed Mean",
    "Standard Deviation",
    "Variance",
    "Minimum",
    "Q1",
    "Q3",
    "Maximum",
    "Range",
    "Interquartile Range",
    "Coefficient of Variation",
    "Skewness",
    "Kurtosis",
    "Observations"
  ),

  AMD = c(
    mean(returns_tbl$amd_return, na.rm = TRUE),
    median(returns_tbl$amd_return, na.rm = TRUE),
    mean(returns_tbl$amd_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE),
    var(returns_tbl$amd_return, na.rm = TRUE),
    min(returns_tbl$amd_return, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.75, na.rm = TRUE),
    max(returns_tbl$amd_return, na.rm = TRUE),
    diff(range(returns_tbl$amd_return, na.rm = TRUE)),
    IQR(returns_tbl$amd_return, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE) /
      mean(returns_tbl$amd_return, na.rm = TRUE),
    skewness(returns_tbl$amd_return, na.rm = TRUE),
    kurtosis(returns_tbl$amd_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$amd_return))
  ),
  
  ETH = c(
    mean(returns_tbl$eth_return, na.rm = TRUE),
    median(returns_tbl$eth_return, na.rm = TRUE),
    mean(returns_tbl$eth_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE),
    var(returns_tbl$eth_return, na.rm = TRUE),
    min(returns_tbl$eth_return, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.75, na.rm = TRUE),
    max(returns_tbl$eth_return, na.rm = TRUE),
    diff(range(returns_tbl$eth_return, na.rm = TRUE)),
    IQR(returns_tbl$eth_return, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE) /
      mean(returns_tbl$eth_return, na.rm = TRUE),
    skewness(returns_tbl$eth_return, na.rm = TRUE),
    kurtosis(returns_tbl$eth_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$eth_return))
  ),
  
  PHI = c(
    mean(returns_tbl$phi_return, na.rm = TRUE),
    median(returns_tbl$phi_return, na.rm = TRUE),
    mean(returns_tbl$phi_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE),
    var(returns_tbl$phi_return, na.rm = TRUE),
    min(returns_tbl$phi_return, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.75, na.rm = TRUE),
    max(returns_tbl$phi_return, na.rm = TRUE),
    diff(range(returns_tbl$phi_return, na.rm = TRUE)),
    IQR(returns_tbl$phi_return, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE) /
      mean(returns_tbl$phi_return, na.rm = TRUE),
    skewness(returns_tbl$phi_return, na.rm = TRUE),
    kurtosis(returns_tbl$phi_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$phi_return))
  ),
  
  SPY = c(
    mean(returns_tbl$phi_return, na.rm = TRUE),
    median(returns_tbl$phi_return, na.rm = TRUE),
    mean(returns_tbl$phi_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE),
    var(returns_tbl$phi_return, na.rm = TRUE),
    min(returns_tbl$phi_return, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.75, na.rm = TRUE),
    max(returns_tbl$phi_return, na.rm = TRUE),
    diff(range(returns_tbl$phi_return, na.rm = TRUE)),
    IQR(returns_tbl$phi_return, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE) /
      mean(returns_tbl$phi_return, na.rm = TRUE),
    skewness(returns_tbl$phi_return, na.rm = TRUE),
    kurtosis(returns_tbl$phi_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$phi_return))
  )
)

descriptive_table
```
# Visualization
"Produce at least eight professional figures with interpretations." -- from the specs

```{r}
```

(for these sections below im not sure if this should be in the paper or in the RMD file itself, or we place it here then elaborate further in the paper)
# Exploratory Financial Analysis
"Compare returns, volatility, inflationary periods, high-VIX periods and interest-rate environments." -- from the specs
```{r}
```

# Investment Recommendation
"Recommend allocation of USD 1,000,000 among the selected U.S. asset, Philippine asset, cryptocurrency and cash." -- from the specs
```{r}
```

# References
- booyah