---
title: "FINSTAD Financial Database"
output: html_document
date: "2026-06-29"
---

# FINSTAD Financial Database

Group 1

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is an integrated database containing various explorations and visualizations of data from a Philippine asset, a U.S. asset, and cryptocurrency, against FRED macroeconomic data and the SPDR S&P 500 ETF Trust. The data analyzed is restricted to January 1, 2020 until December 31, 2025.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, timetk)

library(dplyr)
library(ggplot2)
library(quantmod)
library(tidyverse)
library(tidyquant)
library(lubridate)
library(timetk)
```

### Comparing the different datasets
note: we probably need to explicitly write down and explain each set smt like that

We first get symbols from the following sources:
- FRED
- SPY
- Cryptocurrency
- U.S. asset
- Philippine asset

```{r}
# date constants for use throughout the project.
DATE_FROM <- "2020-01-01"
DATE_TO <- "2025-12-31"

# macroeconomic data is sourced from the Federal Reserve Bank of St. Louis FRED dataset.
quantmod::getSymbols("DGS10",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("CPIAUCSL",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("FEDFUNDS",
                     src = "FRED",
                     from = DATE_FROM,
                     to = DATE_TO)

# the following symbol data is sourced from yahoo finance.
quantmod::getSymbols("SPY",                 # SPDR S&P 500 ETF Trust
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("AMD",                 # U.S. Asset: Advanced Micro Devices, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("PHI",                 # Philippine Asset: PLDT, Inc.
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("^VIX",                # CBOE Volatility Index
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
quantmod::getSymbols("ETH-USD",             # Cryptocurrency: Ethereum
                     src = "yahoo",
                     from = DATE_FROM,
                     to = DATE_TO)
```

Then we can look at the macroeconomic data,

The data is taken from the Federal Reserve Economic Database (FRED) from the Federal Reserve Bank of St. Louis, where we load in the _Market Yield on U.S. Treasury Securities at 10-Year Constant Maturity_ (**DGS10**), _Consumer Price Index for All Urban Consumers_ (**CPIAUCSL**), and the _Federal Funds Effective Rate_ (**FEDFUNDS**).

**DGS10** is used as a proxy for the "risk-free rate" in finance, it helps investors gauge inflation expectations, economic growth, and monetary policy shifts. **CPIAUCSL** is seasonally adjusted to gauge monthly inflation metrics on consumer goods. **FEDFUNDS** is the monthly average interest rate at which banks lend reserves to each other overnight, and reflects the stance of U.S. monetary policy (tightening vs. easing cycles).

These pieces of macroeconomic data would help in understanding broader market cycles and how systemic risks affect stock valuations.
```{r}
head(DGS10)
head(CPIAUCSL)
head(FEDFUNDS)
```

We can then take a look at our chosen financial market data,

U.S. Asset - Advanced Micro Devices, Inc. (AMD)
Philippine Asset - PLDT, Inc.(PHI)
Cryptocurrency - Etherium (ETH or ETH-USD)
Required - State Street SPDR S&P 500 ETF Trust (SPY)

*note: do we explain why we chose these companies lol??
```{r}
ETH <- `ETH-USD`

head(AMD)
head(PHI)
head(ETH)
head(SPY)
```

Lastly, we take a look at the _Cboe Volatility Index_ (**VIX**), this is not an individual stock but a real time market index created by the Chicago Board Options Exchange (Cboe). This measures the market's expectation of 30-day volatility in the S&P 500 index.

This adds aid on top of macroeconomic data to further gauge systemic risks, VIX is known as the "Fear Gauge" which reflects levels of anxiety, complacency, or uncertainty in the market, when this "Fear Gauge" level rises it implies larger expected swings in stock prices.
```{r}
head(VIX)
```

# Cleaning

### 1. Converting XTS to tibbles
First we convert all XTS objects to tibbles for easier data manipulation.
```{r}
dgs10_tbl <- tibble(
  date = index(DGS10),
  dgs10 = as.numeric(DGS10$DGS10),
)
cpiaucsl_tbl <- tibble(
  date = index(CPIAUCSL),
  cpiaucsl = as.numeric(CPIAUCSL$CPIAUCSL),
)
spy_tbl <- tibble(
  date = index(SPY),
  open = as.numeric(SPY$SPY.Open),
  high = as.numeric(SPY$SPY.High),
  low = as.numeric(SPY$SPY.Low),
  close = as.numeric(SPY$SPY.Close),
  volume = as.numeric(SPY$SPY.Volume),
  adjusted = as.numeric(SPY$SPY.Adjusted),
  adj_close = as.numeric(Ad(SPY)),
)
amd_tbl <- tibble(
  date = index(AMD),
  open = as.numeric(AMD$AMD.Open),
  high = as.numeric(AMD$AMD.High),
  low = as.numeric(AMD$AMD.Low),
  close = as.numeric(AMD$AMD.Close),
  volume = as.numeric(AMD$AMD.Volume),
  adjusted = as.numeric(AMD$AMD.Adjusted),
  adj_close = as.numeric(Ad(AMD)),
)
phi_tbl <- tibble(
  date = index(PHI),
  open = as.numeric(PHI$PHI.Open),
  high = as.numeric(PHI$PHI.High),
  low = as.numeric(PHI$PHI.Low),
  close = as.numeric(PHI$PHI.Close),
  volume = as.numeric(PHI$PHI.Volume),
  adjusted = as.numeric(PHI$PHI.Adjusted),
  adj_close = as.numeric(Ad(PHI)),
)
vix_tbl <- tibble(
  date = index(VIX),
  open = as.numeric(VIX$VIX.Open),
  high = as.numeric(VIX$VIX.High),
  low = as.numeric(VIX$VIX.Low),
  close = as.numeric(VIX$VIX.Close),
  volume = as.numeric(VIX$VIX.Volume),
  adjusted = as.numeric(VIX$VIX.Adjusted),
  adj_close = as.numeric(Ad(VIX)),
)
eth_tbl <- tibble(
  date = index(ETH),
  open = as.numeric(ETH$`ETH-USD.Open`),
  high = as.numeric(ETH$`ETH-USD.High`),
  low = as.numeric(ETH$`ETH-USD.Low`),
  close = as.numeric(ETH$`ETH-USD.Close`),
  volume = as.numeric(ETH$`ETH-USD.Volume`),
  adjusted = as.numeric(ETH$`ETH-USD.Adjusted`),
  adj_close = as.numeric(Ad(ETH)),
)
```

### 2. Asset Identifier
We add an asset identifier per asset to see the data better in table structures.
```{r}
amd_tbl <- amd_tbl |> 
  mutate(asset = "AMD")

eth_tbl <- eth_tbl |> 
  mutate(asset = "ETH")

phi_tbl <- phi_tbl |> 
  mutate(asset = "PHI")

spy_tbl <- spy_tbl |> 
  mutate(asset = "SPY")

vix_tbl <- vix_tbl |> 
  mutate(asset = "VIX")
```

### 3. Computing daily returns
idk if we should compute basic returns here or log returns

Daily returns are computed using the basic return formula
```{r}
amd_tbl <- amd_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

eth_tbl <- eth_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

phi_tbl <- phi_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

spy_tbl <- spy_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )

vix_tbl <- vix_tbl |> 
  mutate(
    return = adj_close / lag(adj_close) - 1
  )
```

### 4. Removing missing rows
idk if we should impute or just remove null rows 
```{r}
amd_tbl <- amd_tbl  |> 
  filter(!is.na(return))

eth_tbl <- eth_tbl  |> 
  filter(!is.na(return))

phi_tbl <- phi_tbl |> 
  filter(!is.na(return))

spy_tbl <- spy_tbl  |> 
  filter(!is.na(return))

vix_tbl <- vix_tbl  |> 
  filter(!is.na(return))
```

### 5. Examine the data
```{r}
glimpse(amd_tbl)
glimpse(eth_tbl)
glimpse(phi_tbl)
glimpse(spy_tbl)
glimpse(vix_tbl)
```

Summary Tables
```{r}
summary(amd_tbl)
summary(eth_tbl)
summary(phi_tbl)
summary(spy_tbl)
summary(vix_tbl)
```

### 6. Preparing Macroeconomic variables
```{r}
treasury_tbl <- tibble(
  date = index(DGS10),
  treasury_10y = as.numeric(DGS10$DGS10)
)
```

```{r}
cpi_tbl <- tibble(
  date = index(CPIAUCSL),
  cpi = as.numeric(CPIAUCSL$CPIAUCSL)
)
```

```{r}
fedfunds_tbl <- tibble(
  date = index(FEDFUNDS),
  fedfunds = as.numeric(FEDFUNDS$FEDFUNDS)
)
```

```{r}
head(treasury_tbl)
head(cpi_tbl)
head(fedfunds_tbl)
```

### 7. Create an Observation Summary Table and saving it
We first count the observations
```{r}
nrow(amd_tbl)
nrow(eth_tbl)
nrow(phi_tbl)
nrow(spy_tbl)
nrow(vix_tbl)
nrow(treasury_tbl)
nrow(cpi_tbl)
nrow(fedfunds_tbl)
```

Creating the observation summary table
```{r}
obs_summary <- tibble(
  Dataset = c("AMD",
              "Ethereum",
              "PLDT",
              "SPY",
              "VIX",
              "Treasury",
              "CPI",
              "Fed Funds Rate"),
  Observations = c(
    nrow(amd_tbl),
    nrow(eth_tbl),
    nrow(phi_tbl),
    nrow(spy_tbl),
    nrow(vix_tbl),
    nrow(treasury_tbl),
    nrow(cpi_tbl),
    nrow(fedfunds_tbl)
  )
)

obs_summary
```

Saving the clean tables
```{r}
amd_clean <- amd_tbl
eth_clean <- eth_tbl
phi_clean <- phi_tbl
spy_clean <- spy_tbl
vix_clean <- vix_tbl
treasury_clean <- treasury_tbl
cpi_clean <- cpi_tbl
fedfunds_clean <- fedfunds_tbl
```

# Database Integration
"Explain the SQL join strategy used... Justify the selected join strategy and explain any reduction in observations." 

After cleaning all the data, we collate the rows into 1 unified dataset.

## 1. Renaming variables
```{r}
amd_sql <- amd_clean  |> 
  select(date,
         amd_price = adj_close,
         amd_return = return)

eth_sql <- eth_clean  |> 
  select(date,
         eth_price = adj_close,
         eth_return = return)

phi_sql <- phi_clean  |> 
  select(date,
         phi_price = adj_close,
         phi_return = return)

spy_sql <- spy_clean  |> 
  select(date,
         spy_price = adj_close,
         spy_return = return)

vix_sql <- vix_clean |> 
  select(date,
         vix = adj_close,
         vix_return = return)
```

```{r}
head(amd_sql)
head(eth_sql)
head(phi_sql)
head(spy_sql)
head(vix_sql)
```

## 2. The Trading Calendar Problem

Our four core assets do not trade on the same schedule:

- **SPY** and **AMD** follow the NYSE/NASDAQ calendar (U.S. weekdays, minus U.S. market holidays).
- **PHI** (PLDT ADR) also trades on a U.S. exchange (NYSE), so it largely follows the same calendar as SPY/AMD, but can still diverge on days one market is closed and the other isnt (e.g. differing holiday schedules).
- **ETH** trades every single calendar day, including weekends and all holidays, since crypto markets never close.
- **VIX** follows the CBOE/NYSE calendar, similar to SPY.

This means a naive join on `date` will not line up perfectly, ETH alone contributes roughly 30% more calendar dates per year than the equity assets. This mismatch is exactly why the join strategy matters, and why we need to demonstrate multiple join types before picking one.

## 3. Demonstrating Join Types

Using `spy_sql` (equity calendar) and `eth_sql` (7-day calendar) as the clearest illustration of the mismatch:

```{r}
cat("SPY rows: ", nrow(spy_sql), "\n")
cat("ETH rows: ", nrow(eth_sql), "\n")
```

**`inner_join()`** , keeps only dates present in *both* datasets. This throws away every weekend/holiday ETH observation that SPY doesnt have.
```{r}
demo_inner <- spy_sql |> 
  inner_join(eth_sql, by = "date")

cat("inner_join rows: ", nrow(demo_inner), "\n")
```

**`full_join()`** , keeps every date present in *either* dataset, filling the missing side with `NA`. This preserves ETHs weekend rows but introduces `NA` for SPY/AMD/PHI on those same days, since equities simply didnt trade.
```{r}
demo_full <- spy_sql |> 
  full_join(eth_sql, by = "date")

cat("full_join rows: ", nrow(demo_full), "\n")
cat("Rows where SPY is NA (i.e. weekend/holiday-only ETH dates): ",
    sum(is.na(demo_full$spy_price)), "\n")
```

**`anti_join()`** , shows *only* the rows that dont have a match, which is useful for auditing exactly which dates get dropped or orphaned by a join. Here, it shows every date ETH traded that SPY didnt:
```{r}
demo_anti <- eth_sql |> 
  anti_join(spy_sql, by = "date")

cat("Dates ETH traded but SPY did not: ", nrow(demo_anti), "\n")
head(demo_anti)
```

**`left_join()`** , anchors on the left-hand tables dates and pulls in matches from the right, discarding right-only rows. Anchoring on `spy_sql` keeps every SPY trading day and attaches ETH's price on that date where available:
```{r}
demo_left <- spy_sql |> 
  left_join(eth_sql, by = "date")

cat("left_join rows: ", nrow(demo_left), "\n")
cat("Rows where ETH is NA: ", sum(is.na(demo_left$eth_price)), "\n")
```

## 4. Choosing the Final Join Strategy

Our final database is meant to answer an **equity/benchmark-driven investment question** , comparing AMD, PHI, and ETH against SPY as the reference benchmark, alongside macro conditions. Given that framing:

- `inner_join()` is too aggressive , it would silently drop every day where even one asset didnt trade, shrinking the dataset unnecessarily and losing information about days SPY *did* move.
- `full_join()` keeps the most data, but introduces weekend/holiday rows where SPY, AMD, PHI, and VIX are all structurally `NA` (they simply have no observation those days) , this isnt "missing" data to be filled, its a non-event, so those extra rows have no analytical value for an equity-anchored study.
- `anti_join()` is a diagnostic tool, not a database-building join , we use it above to confirm exactly how many ETH-only dates exist, but it doesnt belong in the final chain.
- **`left_join()` anchored on `spy_sql`** is our final choice: it preserves every SPY trading day (our benchmark and reference calendar), attaches whichever other series have data on that date, and cleanly produces `NA` only where a genuine data gap exists (e.g. ETH trading on a day AMD didnt, or a macro series not yet updated) rather than manufacturing rows that dont correspond to any equity trading activity.

## 5. Building the Unified Database
```{r}
unified_db <- spy_sql |> 
  left_join(amd_sql,
            by = "date") |> 
  left_join(phi_sql,
            by = "date") |>
  left_join(eth_sql,
            by = "date") |> 
  left_join(vix_sql,
            by = "date") |> 
  left_join(treasury_clean,
            by = "date") |> 
  left_join(cpi_clean,
            by = "date") |> 
  left_join(fedfunds_clean,
            by = "date") |> 
  arrange(date) |>
  tidyr::fill(treasury_10y,
              cpi,
              fedfunds,
              .direction = "down")

glimpse(unified_db)
head(unified_db)
view(unified_db)
```

`treasury_10y` and `fedfunds` are only reported on U.S. business days, while `cpi` is only reported once a month. Since our final database is anchored on SPY's daily trading calendar (via `left_join()`), joining these lower-frequency series directly leaves `NA` on every date without a fresh reading. We resolve this with `tidyr::fill(..., .direction = "down")`, which carries the most recent known value forward , e.g. a CPI print from early June applies to every trading day until the next print in early July. This is standard practice for macro series in daily financial datasets, since the "current" CPI or Fed Funds level doesnt change on the days between official releases.

## 6. Observation Tracking

Tracking row counts at each stage of the join chain shows exactly where (and why) the dataset shrinks or holds steady.
```{r}
obs_track_spy    <- spy_sql
obs_track_amd    <- obs_track_spy |> left_join(amd_sql, by = "date")
obs_track_phi    <- obs_track_amd |> left_join(phi_sql, by = "date")
obs_track_eth    <- obs_track_phi |> left_join(eth_sql, by = "date")
obs_track_vix    <- obs_track_eth |> left_join(vix_sql, by = "date")
obs_track_treas  <- obs_track_vix |> left_join(treasury_clean, by = "date")
obs_track_cpi    <- obs_track_treas |> left_join(cpi_clean, by = "date")
obs_track_ff     <- obs_track_cpi |> left_join(fedfunds_clean, by = "date")

join_tracking <- tibble(
  Step = c("Original SPY (anchor)",
           "+ AMD",
           "+ PHI",
           "+ ETH",
           "+ VIX",
           "+ Treasury (DGS10)",
           "+ CPI",
           "+ Fed Funds Rate"),
  Rows_After_Join = c(
    nrow(obs_track_spy),
    nrow(obs_track_amd),
    nrow(obs_track_phi),
    nrow(obs_track_eth),
    nrow(obs_track_vix),
    nrow(obs_track_treas),
    nrow(obs_track_cpi),
    nrow(obs_track_ff)
  )
)

join_tracking
```

Because every join in the chain is a `left_join()` anchored on `spy_sql`, **row count never shrinks past the original SPY row count** , a `left_join()` can only add columns/matches or introduce `NA`, never drop rows from the left-hand side. This is exactly the guarantee we wanted from our join strategy: the equity benchmark's trading calendar is preserved intact all the way through.

Checking null rows after the join and fill step:
```{r}
colSums(
  is.na(
    unified_db
  )
)
```

## 7. Dropping Null Rows

After `tidyr::fill()`, the macro columns (`treasury_10y`, `cpi`, `fedfunds`) should have no remaining `NA` past their first reported observation. Any `NA` still present at this point belongs to the **asset price/return columns** (`amd_price`, `amd_return`, `phi_price`, `phi_return`, `eth_price`, `eth_return`, `vix`, `vix_return`) , these represent real gaps where a given asset simply didnt trade on a day SPY did (e.g. a PSE holiday that isnt a NYSE holiday), not something we can reasonably forward-fill without fabricating price movement that never happened.

We drop these remaining rows, since imputing a stock/crypto price would distort the return and volatility statistics that our entire analysis depends on:
```{r}
unified_clean <- unified_db |> 
  tidyr::drop_na(amd_price, amd_return,
                 phi_price, phi_return,
                 eth_price, eth_return,
                 vix, vix_return,
                 spy_price, spy_return)

cat("Rows before dropping asset-level NA: ", nrow(unified_db), "\n")
cat("Rows after dropping asset-level NA: ", nrow(unified_clean), "\n")
cat("Rows dropped: ", nrow(unified_db) - nrow(unified_clean), "\n")

colSums(is.na(unified_clean))
```

`unified_clean` is our final analytical dataset, used for all descriptive statistics, SQL investigations, and visualizations going forward.

# SQL Investigation
"Investigate at least 10 meaningful financial questions using select(), filter(), mutate(), arrange(), group_by() and summarise(). Each query must include Objective, Code, Output and Interpretation." 

## Investigation 1: Biggest single-day moves per asset

**Objective:** Find the top 3 single-day returns for each of the four core assets to see which asset produces the largest short-term shocks.

**Code:**
```{r}
returns_long <- unified_clean |>
  select(date, amd_return, eth_return, phi_return, spy_return) |>
  tidyr::pivot_longer(cols = -date, names_to = "asset", values_to = "return")

top_gains <- returns_long |>
  group_by(asset) |>
  arrange(desc(return), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  filter(rank <= 3) |>
  select(asset, date, return, rank)

top_gains
```

**Output:** *(run the chunk above and paste the printed table into your report)*

**Interpretation:** *After running this, compare which asset's top return is the largest in magnitude. ETH will almost certainly show the biggest single-day swings given crypto's higher baseline volatility, while SPY should show the smallest. Note the actual dates too, since clustering around the same few dates suggests a shared market-wide shock rather than an asset-specific event.*

## Investigation 2: Monthly average returns

**Objective:** Compute the average daily return per calendar month for each asset to identify which months or periods were consistently strong or weak.

**Code:**
```{r}
monthly_avg_returns <- unified_clean |>
  mutate(year_month = format(date, "%Y-%m")) |>
  group_by(year_month) |>
  summarise(
    avg_amd = mean(amd_return, na.rm = TRUE),
    avg_eth = mean(eth_return, na.rm = TRUE),
    avg_phi = mean(phi_return, na.rm = TRUE),
    avg_spy = mean(spy_return, na.rm = TRUE)
  )

head(monthly_avg_returns, 10)
```



**Interpretation:** *Look for months where all four columns are negative together, this points to broad market stress rather than an asset-specific dip. Months where only ETH or only PHI moves against the others suggest asset-specific news rather than macro conditions.*

## Investigation 3: SPY selloff days and cross-asset behavior

**Objective:** Identify SPY's 10 worst trading days and check how AMD, PHI, and ETH performed on those same dates, to test whether any of them behave as a diversifier during U.S. market selloffs.

**Code:**
```{r}
spy_selloffs <- unified_clean |>
  filter(spy_return < 0) |>
  arrange(spy_return) |>
  select(date, spy_return, amd_return, phi_return, eth_return) |>
  head(10)

spy_selloffs
```



**Interpretation:** *If AMD and ETH are also deeply negative on SPY's worst days, that is evidence they move with the U.S. market rather than against it, meaning they add little diversification value. If PHI holds up better on these dates, that supports including it as a partial hedge in the final allocation.*

## Investigation 4: High-VIX vs low-VIX return and volatility comparison

**Objective:** Split trading days into High VIX and Low VIX regimes (relative to the median VIX level) and compare average return and volatility in each regime.

**Code:**
```{r}
vix_median <- median(unified_clean$vix, na.rm = TRUE)

vix_regime_stats <- unified_clean |>
  mutate(vix_regime = if_else(vix > vix_median, "High VIX", "Low VIX")) |>
  group_by(vix_regime) |>
  summarise(
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    sd_spy_return = sd(spy_return, na.rm = TRUE),
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    sd_amd_return = sd(amd_return, na.rm = TRUE),
    n_days = n()
  )

vix_regime_stats
```



**Interpretation:** *Expect the standard deviation columns to be visibly higher in the High VIX regime, confirming VIX is doing its job as a "fear gauge." If average returns are also lower (or negative) in the High VIX regime, that supports treating elevated VIX as a signal to reduce exposure or hold more cash.*

## Investigation 5: Returns across Fed Funds rate regimes

**Objective:** Classify each trading day by whether the Fed Funds Rate was rising, falling, or holding steady, then compare average returns across the three regimes to gauge rate sensitivity.

**Code:**
```{r}
fedfunds_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    fedfunds_change = fedfunds - lag(fedfunds),
    rate_regime = case_when(
      fedfunds_change > 0 ~ "Rate Hike",
      fedfunds_change < 0 ~ "Rate Cut",
      TRUE ~ "Held Steady"
    )
  ) |>
  filter(!is.na(rate_regime)) |>
  group_by(rate_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

fedfunds_trend
```



**Interpretation:** *Crypto (ETH) is typically the most rate-sensitive asset class since it has no yield of its own to compete with rising rates. If ETH's average return is clearly weakest during Rate Hike days relative to the others, that is useful evidence for the investment recommendation section, especially if the current environment is a hiking or holding cycle.*

## Investigation 6: Returns across inflation regimes

**Objective:** Classify each trading day by whether CPI was rising or falling/flat that month, then compare average returns across the two inflation regimes.

**Code:**
```{r}
cpi_trend <- unified_clean |>
  arrange(date) |>
  mutate(
    cpi_change = cpi - lag(cpi),
    inflation_regime = if_else(cpi_change > 0, "CPI Rising", "CPI Falling or Flat")
  ) |>
  filter(!is.na(inflation_regime)) |>
  group_by(inflation_regime) |>
  summarise(
    avg_amd_return = mean(amd_return, na.rm = TRUE),
    avg_eth_return = mean(eth_return, na.rm = TRUE),
    avg_phi_return = mean(phi_return, na.rm = TRUE),
    avg_spy_return = mean(spy_return, na.rm = TRUE),
    n_days = n()
  )

cpi_trend
```



**Interpretation:** *If returns are noticeably weaker during CPI Rising periods across most assets, that supports the idea that inflation surprises pressure both equities and crypto. Since CPI only updates monthly and we forward-filled it, keep in mind this regime label applies to the whole month following a print, not just the release day itself.*

## Investigation 7: Co-movement across all four assets

**Objective:** Count how often all four assets move in the same direction on the same day (all up or all down), to measure how correlated the portfolio really is versus how often the assets diverge.

**Code:**
```{r}
co_movement <- unified_clean |>
  mutate(
    all_positive = amd_return > 0 & eth_return > 0 & phi_return > 0 & spy_return > 0,
    all_negative = amd_return < 0 & eth_return < 0 & phi_return < 0 & spy_return < 0
  ) |>
  summarise(
    all_up_days = sum(all_positive, na.rm = TRUE),
    all_down_days = sum(all_negative, na.rm = TRUE),
    total_days = n(),
    pct_co_moving = (sum(all_positive, na.rm = TRUE) + sum(all_negative, na.rm = TRUE)) / n() * 100
  )

co_movement
```



**Interpretation:** *A high `pct_co_moving` value means the four assets rarely offset each other, which weakens the diversification case for holding all of them together. A lower value means there is meaningful room for one asset to cushion losses in another, which is a stronger argument for spreading the USD 1,000,000 allocation across all three risk assets instead of concentrating it.*

## Investigation 8: Highest-volume trading days for AMD and ETH

**Objective:** Identify the 5 highest-volume trading days for AMD and ETH individually, since volume spikes often coincide with earnings releases, major news, or sharp price moves.

**Code:**
```{r}
amd_top_volume <- amd_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

eth_top_volume <- eth_clean |>
  select(date, volume, return) |>
  arrange(desc(volume)) |>
  head(5)

amd_top_volume
eth_top_volume
```



**Interpretation:** *Check whether the high-volume dates for each asset line up with large positive or negative returns. If a volume spike happens without a big price move, that could indicate a technical event (index rebalancing, options expiry) rather than fresh information entering the market.*

## Investigation 9: Quarterly average price levels

**Objective:** Compute the average closing price per asset by quarter to see the longer-term price trend without daily noise.

**Code:**
```{r}
quarterly_prices <- unified_clean |>
  mutate(year_quarter = paste0(lubridate::year(date), "-Q", lubridate::quarter(date))) |>
  group_by(year_quarter) |>
  summarise(
    avg_spy_price = mean(spy_price, na.rm = TRUE),
    avg_amd_price = mean(amd_price, na.rm = TRUE),
    avg_phi_price = mean(phi_price, na.rm = TRUE),
    avg_eth_price = mean(eth_price, na.rm = TRUE)
  )

quarterly_prices
```



**Interpretation:** *This table is the basis for your long-term price trend chart in the Visualization section. Look at which asset has the steadiest quarter-over-quarter climb versus which one swings the most, since that speaks directly to how much of the allocation should go to growth versus stability.*

## Investigation 10: Risk-adjusted return comparison

**Objective:** Compute a simple Sharpe-style ratio for each asset (average daily return minus a daily risk-free rate, divided by return standard deviation) using the 10-Year Treasury yield as the risk-free proxy, to compare returns on a risk-adjusted basis rather than raw returns alone.

**Code:**
```{r}
risk_free_daily <- mean(unified_clean$treasury_10y, na.rm = TRUE) / 100 / 252

risk_adjusted <- unified_clean |>
  select(amd_return, eth_return, phi_return, spy_return) |>
  summarise(
    amd_sharpe = (mean(amd_return, na.rm = TRUE) - risk_free_daily) / sd(amd_return, na.rm = TRUE),
    eth_sharpe = (mean(eth_return, na.rm = TRUE) - risk_free_daily) / sd(eth_return, na.rm = TRUE),
    phi_sharpe = (mean(phi_return, na.rm = TRUE) - risk_free_daily) / sd(phi_return, na.rm = TRUE),
    spy_sharpe = (mean(spy_return, na.rm = TRUE) - risk_free_daily) / sd(spy_return, na.rm = TRUE)
  )

risk_adjusted
```



**Interpretation:** *This is the single most important table for the Investment Recommendation section. The asset with the highest Sharpe-style ratio delivered the best return per unit of risk taken, not just the highest raw return. A high-return asset with an even higher volatility can still end up with a worse ratio than a steadier asset, so use this table, not just the raw mean returns from Investigation 2, to justify how you split the USD 1,000,000 allocation.*

# Descriptive Statistics
"Compute mean, median, variance, standard deviation, minimum, maximum, range, skewness and kurtosis." 

From the previous section, a unified database named `unified_clean` was created , this is our final analytical dataset (post-join, post-fill, post-drop) that all statistics below are computed from.
```{r}
glimpse(unified_clean)
dim(unified_clean)
```

### 1. Selecting return variables
Since our end goal is an investment recommendation, we focus this analysis on the daily return series for each tradable asset (return = risk/reward), while VIX and macro variables are analyzed separately as context in the Exploratory Financial Analysis section.
```{r}
returns_tbl <- unified_clean  |> 

  select(date,
         amd_return,
         eth_return,
         phi_return,
         spy_return)

head(returns_tbl)
dim(returns_tbl)

```

### 2. Computing median and mean
```{r}
#Mean Return
cat("AMD Mean: ")
mean(returns_tbl$amd_return,
     na.rm = TRUE)

#Median Return
cat("AMD Median: ")
median(returns_tbl$amd_return,
       na.rm = TRUE)

#Trimmed Mean
cat("AMD Trimmed Mean: ")
mean(returns_tbl$amd_return,
     trim = 0.05,
     na.rm = TRUE)

cat("ETH Mean: ")
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("ETH Median: ")
median(returns_tbl$eth_return,
       na.rm = TRUE)

cat("ETH Trimmed Mean: ")
mean(returns_tbl$eth_return,
     trim = 0.05,
     na.rm = TRUE)

cat("PHI Mean: ")
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("PHI Median: ")
median(returns_tbl$phi_return,
       na.rm = TRUE)

cat("PHI Trimmed Mean: ")
mean(returns_tbl$phi_return,
     trim = 0.05,
     na.rm = TRUE)

cat("SPY Mean: ")
mean(returns_tbl$spy_return,
     na.rm = TRUE)

cat("SPY Median: ")
median(returns_tbl$spy_return,
       na.rm = TRUE)

cat("SPY Trimmed Mean: ")
mean(returns_tbl$spy_return,
     trim = 0.05,
     na.rm = TRUE)

```

### 3. Computing Dispersion (Risk)
variance, std dev, range, interquartile range, coefficient of variation
```{r}
#Variance
cat("AMD Variance: ")
var(returns_tbl$amd_return,
    na.rm = TRUE)

#Standard Deviation
cat("AMD Standard Deviation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE)

#Range
cat("AMD Range: ")
range(returns_tbl$amd_return,
      na.rm = TRUE)

#Interquartile Range
cat("AMD IQR: ")
IQR(returns_tbl$amd_return,
    na.rm = TRUE)

#Coefficient of Variation
cat("AMD Coeff. of Variation: ")
sd(returns_tbl$amd_return,
   na.rm = TRUE) /
mean(returns_tbl$amd_return,
     na.rm = TRUE)

cat("ETH Variance: ")
var(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Standard Deviation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE)

cat("ETH Range: ")
range(returns_tbl$eth_return,
      na.rm = TRUE)

cat("ETH IQR: ")
IQR(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Coeff. of Variation: ")
sd(returns_tbl$eth_return,
   na.rm = TRUE) /
mean(returns_tbl$eth_return,
     na.rm = TRUE)

cat("PHI Variance: ")
var(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Standard Deviation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE)

cat("PHI Range: ")
range(returns_tbl$phi_return,
      na.rm = TRUE)

cat("PHI IQR: ")
IQR(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Coeff. of Variation: ")
sd(returns_tbl$phi_return,
   na.rm = TRUE) /
mean(returns_tbl$phi_return,
     na.rm = TRUE)

cat("SPY Variance: ")
var(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Standard Deviation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE)

cat("SPY Range: ")
range(returns_tbl$spy_return,
      na.rm = TRUE)

cat("SPY IQR: ")
IQR(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Coeff. of Variation: ")
sd(returns_tbl$spy_return,
   na.rm = TRUE) /
mean(returns_tbl$spy_return,
     na.rm = TRUE)

```

### 4. Viewing data shape
skewness and kurtosis
```{r}
library(moments)
#Skewness
cat("AMD Skewness: ")
skewness(returns_tbl$amd_return,
         na.rm = TRUE)
#Kurtosis
cat("AMD Kurtosis: ")
kurtosis(returns_tbl$amd_return,
         na.rm = TRUE)

cat("ETH Skewness: ")
skewness(returns_tbl$eth_return,
         na.rm = TRUE)

cat("ETH Kurtosis: ")
kurtosis(returns_tbl$eth_return,
         na.rm = TRUE)

cat("PHI Skewness: ")
skewness(returns_tbl$phi_return,
         na.rm = TRUE)

cat("PHI Kurtosis: ")
kurtosis(returns_tbl$phi_return,
         na.rm = TRUE)

cat("SPY Skewness: ")
skewness(returns_tbl$spy_return,
         na.rm = TRUE)

cat("SPY Kurtosis: ")
kurtosis(returns_tbl$spy_return,
         na.rm = TRUE)
```

### 5. Computing data position ?
min, max, quartiles, percentiles
```{r}
#Minimum
cat("AMD Minimum: ")
min(returns_tbl$amd_return,
    na.rm = TRUE)

#Maximum
cat("AMD Maximum: ")
max(returns_tbl$amd_return,
    na.rm = TRUE)

#Quartiles
cat("AMD Quartiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

#Percentiles
cat("AMD Percentiles: ")
quantile(returns_tbl$amd_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("ETH Minimum: ")
min(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Maximum: ")
max(returns_tbl$eth_return,
    na.rm = TRUE)

cat("ETH Quartiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("ETH Percentiles: ")
quantile(returns_tbl$eth_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("PHI Minimum: ")
min(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Maximum: ")
max(returns_tbl$phi_return,
    na.rm = TRUE)

cat("PHI Quartiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("PHI Percentiles: ")
quantile(returns_tbl$phi_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)

cat("SPY Minimum: ")
min(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Maximum: ")
max(returns_tbl$spy_return,
    na.rm = TRUE)

cat("SPY Quartiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.25,.50,.75),
         na.rm = TRUE)

cat("SPY Percentiles: ")
quantile(returns_tbl$spy_return,
         probs = c(.01,.05,.95,.99),
         na.rm = TRUE)
```

### 6. Dealing with Missing Values
```{r}
colSums(is.na(returns_tbl))
```

### 7. Correlation Matrix
```{r}
correlation_matrix <- returns_tbl  |> 
  select(-date)  |> 
  cor(use = "complete.obs")
correlation_matrix
```

### 8. Covariance Matrix 
```{r}
covariance_matrix <- returns_tbl |> 
  select(-date)  |> 
  cov(use = "complete.obs")
covariance_matrix
```

### 9. Final Descriptive Statistics Table
```{r}
library(psych)

desc_stats <- psych::describe(
  returns_tbl  |> 
    select(-date)
)
round(desc_stats,4)
```

```{r}
descriptive_table <- tibble(

  Statistic = c(
    "Mean",
    "Median",
    "Trimmed Mean",
    "Standard Deviation",
    "Variance",
    "Minimum",
    "Q1",
    "Q3",
    "Maximum",
    "Range",
    "Interquartile Range",
    "Coefficient of Variation",
    "Skewness",
    "Kurtosis",
    "Observations"
  ),

  AMD = c(
    mean(returns_tbl$amd_return, na.rm = TRUE),
    median(returns_tbl$amd_return, na.rm = TRUE),
    mean(returns_tbl$amd_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE),
    var(returns_tbl$amd_return, na.rm = TRUE),
    min(returns_tbl$amd_return, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$amd_return, 0.75, na.rm = TRUE),
    max(returns_tbl$amd_return, na.rm = TRUE),
    diff(range(returns_tbl$amd_return, na.rm = TRUE)),
    IQR(returns_tbl$amd_return, na.rm = TRUE),
    sd(returns_tbl$amd_return, na.rm = TRUE) /
      mean(returns_tbl$amd_return, na.rm = TRUE),
    skewness(returns_tbl$amd_return, na.rm = TRUE),
    kurtosis(returns_tbl$amd_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$amd_return))
  ),
  
  ETH = c(
    mean(returns_tbl$eth_return, na.rm = TRUE),
    median(returns_tbl$eth_return, na.rm = TRUE),
    mean(returns_tbl$eth_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE),
    var(returns_tbl$eth_return, na.rm = TRUE),
    min(returns_tbl$eth_return, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$eth_return, 0.75, na.rm = TRUE),
    max(returns_tbl$eth_return, na.rm = TRUE),
    diff(range(returns_tbl$eth_return, na.rm = TRUE)),
    IQR(returns_tbl$eth_return, na.rm = TRUE),
    sd(returns_tbl$eth_return, na.rm = TRUE) /
      mean(returns_tbl$eth_return, na.rm = TRUE),
    skewness(returns_tbl$eth_return, na.rm = TRUE),
    kurtosis(returns_tbl$eth_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$eth_return))
  ),
  
  PHI = c(
    mean(returns_tbl$phi_return, na.rm = TRUE),
    median(returns_tbl$phi_return, na.rm = TRUE),
    mean(returns_tbl$phi_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE),
    var(returns_tbl$phi_return, na.rm = TRUE),
    min(returns_tbl$phi_return, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$phi_return, 0.75, na.rm = TRUE),
    max(returns_tbl$phi_return, na.rm = TRUE),
    diff(range(returns_tbl$phi_return, na.rm = TRUE)),
    IQR(returns_tbl$phi_return, na.rm = TRUE),
    sd(returns_tbl$phi_return, na.rm = TRUE) /
      mean(returns_tbl$phi_return, na.rm = TRUE),
    skewness(returns_tbl$phi_return, na.rm = TRUE),
    kurtosis(returns_tbl$phi_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$phi_return))
  ),
  
  SPY = c(
    mean(returns_tbl$spy_return, na.rm = TRUE),
    median(returns_tbl$spy_return, na.rm = TRUE),
    mean(returns_tbl$spy_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE),
    var(returns_tbl$spy_return, na.rm = TRUE),
    min(returns_tbl$spy_return, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$spy_return, 0.75, na.rm = TRUE),
    max(returns_tbl$spy_return, na.rm = TRUE),
    diff(range(returns_tbl$spy_return, na.rm = TRUE)),
    IQR(returns_tbl$spy_return, na.rm = TRUE),
    sd(returns_tbl$spy_return, na.rm = TRUE) /
      mean(returns_tbl$spy_return, na.rm = TRUE),
    skewness(returns_tbl$spy_return, na.rm = TRUE),
    kurtosis(returns_tbl$spy_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$spy_return))
  )
)

descriptive_table
```
# Visualization
"Produce at least eight professional figures with interpretations." 

```{r}
```

(for these sections below im not sure if this should be in the paper or in the RMD file itself, or we place it here then elaborate further in the paper)
# Exploratory Financial Analysis
"Compare returns, volatility, inflationary periods, high-VIX periods and interest-rate environments." 
```{r}
```

# Investment Recommendation
"Recommend allocation of USD 1,000,000 among the selected U.S. asset, Philippine asset, cryptocurrency and cash." 
```{r}
```

# References
- booyah